---
title: "CL-PFU database pipeline diagnostic report"
output:
  bookdown::pdf_document2:
    toc: false
    citation_package: natbib
author: Matthew Kuperus Heun
date: "`r Sys.Date()`"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(ggpubr)
library(Matrix)
library(matsindf)
library(MKHthemes)
library(tidyr)
```


# Introduction

This report provides diagnostics for the CL-PFU database
pipeline.
The purpose of this report is to compare key results 
from one version of the database to another.


# Preliminaries

We first set the version of the database
for which diagnostics are desired.

```{r version} 
clpfu_version <- "v3.0a2"
```

Set data file paths.

```{r data-file-paths}
world_data_file <- file.path("world_mats.rds")
```

We can download data using this chunk of code.
To save time, 
this chunk is not normally evaluated when building the report
(`eval=FALSE`).
Run this chunk as needed.

```{r download-data, eval=FALSE}
# We first make a connection to the database.
conn <- PFUPipelineTools::get_scratchmdb_conn()
on.exit(DBI::dbDisconnect(conn))

# Next, download desired data and save locally for later use. 
# We do this once to save time.
world_mats <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)

# Save to a file
world_mats |>
  saveRDS(file = world_data_file)

# Disconnect from the database management system.
DBI::dbDisconnect(conn)
```

Load data from the pre-saved file.

```{r load, echo=TRUE}
world_mats <- readRDS(file = world_data_file)
```

At this point, we are ready to make diagnostic graphs and tables.


# Primary, final, and useful aggregations



## Global Final and useful energy and exergy by sector



```{r fig6, fig.width=6, fig.height=3, fig.align='center', fig.retina=2, fig.cap='Remake of Figure 6.', echo=FALSE}


# dplyr::bind_rows(final_aggregates_by_industry |>
#                    dplyr::mutate(stage = "Final"),
#                  useful_aggregates_by_industry |>
#                    dplyr::mutate(stage = "Useful")) |>
#   dplyr::filter(Country == "World") |>
#   ggplot2::ggplot(mapping = ggplot2::aes(x = Year, y = matvals, fill = colnames)) +
#   ggplot2::geom_area() +
#   ggplot2::facet_grid(rows = dplyr::vars(stage), cols = dplyr::vars(Energy.type)) +
#   ggplot2::labs(fill = NULL) +
#   ggplot2::scale_x_continuous(name = NULL, breaks = c(1980, 2000, 2020)) +
#   ggplot2::scale_y_continuous(name = "Energy and exergy [EJ/year]") +
#   ggplot2::scale_fill_manual(values = sector_colours) +
#   MKHthemes::xy_theme() +
#   ggplot2::theme(legend.text = element_text(size = legend_text_size))

```

## Global useful energy and exergy by energy carrier

The following figure is a recreation of Figure 7
of @Brockway:2024ab.

```{r fig7, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross useful energy and exergy consumption and shares by energy product group. Non-energy uses are excluded. This is a recreation (using CL-PFU database ', clpfu_version, ') of Figure 7 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).'), echo=FALSE}
# Define sectors to exclude from the data
# to match sectors excluded from Figure 7.
sectors_to_exclude <- RCLabels::make_or_pattern(
  c("Exports", 
    "Exports to World aviation bunkers", 
    "Exports to World marine bunkers", 
    "Losses", 
    "Statistical differences", 
    "Stock changes"))

# Build the data frame.
grouped_useful_data <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE, # Exclude non-energy uses
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::mutate(
    # For gross useful energy, we need to include both Y and U_EIOU matrices
    YUEIOU = matsbyname::sum_byname(Y, U_EIOU) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)  |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU") |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |>
  dplyr::select(-R, -U, -U_feed, -U_EIOU, -r_EIOU, -V, -Y, -S_units) |>
  tidyr::pivot_longer(cols = "YUEIOU", names_to = "matnames", values_to = "matvals") |>
  
  
  # This code is here only because I forgot to add
  # some heat categories to the eu_product aggregation map.
  # I have now added those aggregations, 
  # so the code below will not be needed in the future.
  # ---MKH, 10 July 2026
  dplyr::mutate(
    matvals = matsbyname::aggregate_byname(
      matvals,
      aggregation_map = list(list(
        HTH = c("HTH", "HTH.1200.C", "HTH.1800.C"), 
        MTH = c("MTH", "MTH.120.C", "MTH.130.C", "MTH.150.C", "MTH.160.C"), 
        LTH = c("LTH", "LTH.90.C")
      )), 
      margin = 1)
  ) |>
  
  matsindf::expand_to_tidy(drop = 0) |>
  # Adjust the names of useful energy carriers to be more readable
  dplyr::mutate(
    rownames = dplyr::case_when(
      rownames == "AnMech" ~ "Animal mechanical\nwork",
      rownames == "HPF" ~ "High-pressure\nfluids",
      rownames == "HTH" ~ "High-temperature\nheat",
      rownames == "HuMech" ~ "Human mechanical\nwork",
      rownames == "IP" ~ "Information\nprocessing",
      rownames == "L" ~ "Light",
      rownames == "LTC" ~ "Low-temperature\ncooling",
      rownames == "LTH" ~ "Low-temperature\nheat",
      rownames == "MechW" ~ "Mechanical work",
      rownames == "MTH" ~ "Medium-temperature\nheat",
      rownames == "NEU" ~ "Non-energy\nuse",
      rownames == "Prop" ~ "Propulsion",
      TRUE ~ .data[["rownames"]]),
    rownames = factor(rownames, levels = c("NEU", "Non-energy\nuse",
                                           "IP", "Information processing", "Information\nprocessing",
                                           "L", "Light",
                                           "HTH", "High-temperature heat", "High-temperature\nheat",
                                           "MTH", "Medium-temperature heat", "Medium-temperature\nheat",
                                           "LTH", "Low-temperature heat", "Low-temperature\nheat",
                                           "LTC", "Low-temperature cooling", "Low-temperature\ncooling",
                                           "HPF", "High-pressure fluids", "High-pressure\nfluids",
                                           "Prop", "Propulsion",
                                           "MechW", "Mechanical work",
                                           "HuMech", "Human mechanical work", "Human mechanical\nwork",
                                           "AnMech", "Animal mechanical work", "Animal mechanical\nwork"))
  )

  
  # dplyr::mutate(
  #   rownames = factor(rownames, levels = c("IP", "L", "HTH", "MTH", "LTH",
  #                                          "LTC", "HPF", "Prop", "MechW",
  #                                          "HuMech", "AnMech"))
  # )

legend_text_size <- 8

# Make the graph
useful_ej <- grouped_useful_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 250, by = 50)) +
  ggplot2::scale_fill_manual(values = MKHthemes::useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme() + 
  ggplot2::theme(legend.text = element_text(size = legend_text_size),
                 axis.text.x = element_blank(),
                 axis.ticks.x = element_blank())


useful_fractions <- grouped_useful_data |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) +
  ggplot2::geom_area(position = ggplot2::position_fill()) +
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Shares [%]",
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  ggplot2::scale_fill_manual(values = MKHthemes::useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size),
                 strip.text.x = ggplot2::element_blank())

ggpubr::ggarrange(useful_ej,
                  useful_fractions,
                  nrow = 2,
                  common.legend = TRUE,
                  legend = "right",
                  align = "v")
```



# References
