---
title: "CL-PFU database pipeline diagnostic report"
output:
  bookdown::pdf_document2:
    toc: false
    citation_package: natbib
author: Matthew Kuperus Heun
date: "`r Sys.Date()`"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(ggpubr)
library(Matrix)
library(matsindf)
library(MKHthemes)
library(tidyr)

legend_text_size <- 8
```


# Introduction

This report provides diagnostics for the CL-PFU database
pipeline.
The purpose of this report is to compare key results 
from one version of the database to another.


# Preliminaries

We first set the version of the database
for which diagnostics are desired.

```{r version} 
clpfu_version <- "v3.0a2"
```

Set data file paths.

```{r data-file-paths}
world_data_file <- file.path("world_mats.rds")
world_eta_file <- file.path("world_eta.rds")
```

We can download data using this chunk of code.
To save time, 
this chunk is not normally evaluated when building the report
(`eval=FALSE`).
Run this chunk as needed.

```{r download-data, eval=FALSE}
# We first make a connection to the database.
conn <- PFUPipelineTools::get_scratchmdb_conn()
on.exit(DBI::dbDisconnect(conn))

# Next, download desired data and save locally for later use. 
# We do this once to save time.
world_mats <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)

# Save to a file
world_mats |>
  saveRDS(file = world_data_file)

world_eta <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "AggEtaPFU", 
  Country == "World", 
  conn = conn, 
  collect = TRUE
)

world_eta |> 
  saveRDS(file = world_eta_file)

# Disconnect from the database management system.
DBI::dbDisconnect(conn)
```

Load data from the pre-saved file.

```{r load, echo=TRUE}
world_mats <- readRDS(file = world_data_file)
world_eta <- readRDS(file = world_eta_file)
```

At this point, we are ready to make diagnostic graphs and tables.


# Primary, final, and useful aggregations



## Global final and useful energy and exergy by sector



```{r fig6, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross final and useful energy and exergy consumption by sector. Non-energy uses are included at the final energy stage. This is a recreation (using CL-PFU database ', clpfu_version, ') of Figure 7 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).'), echo=FALSE}

# Define sectors to exclude from the data
# to match sectors excluded from Figures 6 and 7.
sectors_to_exclude <- RCLabels::make_or_pattern(
  c("Exports", 
    "Exports to World aviation bunkers", 
    "Exports to World marine bunkers", 
    "Losses", 
    "Statistical differences", 
    "Stock changes"))

grouped_final_useful_data <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType, LastStage) |> 
  dplyr::mutate(
    # For gross useful energy, we need to include both Y and U_EIOU matrices
    YUEIOU = matsbyname::sum_byname(Y, U_EIOU) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)  |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU") |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |> 
  dplyr::select(-R, -U, -U_feed, -U_EIOU, -r_EIOU, -V, -Y, -S_units)

grouped_final_useful_data |> 
  dplyr::mutate(
    YUEIOU = matsbyname::colsums_byname(YUEIOU)
  ) |> 
  tidyr::pivot_longer(cols = "YUEIOU", 
                      names_to = "matnames", 
                      values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  dplyr::filter(IncludesNEU == TRUE) |> 
  dplyr::mutate(
    colnames = dplyr::case_when(
      colnames == "Agriculture, forestry, and fishing" ~ "Agriculture, forestry,\nand fishing",
      colnames == "Commercial and public services" ~ "Commercial and\npublic services",
      colnames == "Final consumption not elsewhere specified" ~ "Final consumption not\nelsewhere specified",
      colnames == "Energy industry own use" ~ "Energy industry\nown use",
      TRUE ~ .data[["colnames"]]
    ), 
    colnames = factor(colnames,
                      levels = c("Energy industry\nown use",
                                 "Non-energy use",
                                 "Final consumption not\nelsewhere specified",
                                 "Transport",
                                 "Commercial and\npublic services",
                                 "Industry",
                                 "Residential",
                                 "Agriculture, forestry,\nand fishing"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = colnames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(rows = ggplot2::vars(LastStage), 
                      cols = ggplot2::vars(EnergyType)) + 
  ggplot2::labs(x = NULL, 
                fill = NULL) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Energy and exergy [EJ/year]") +
  ggplot2::scale_fill_manual(values = MKHthemes::sector_colours) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size))
```

## Global useful energy and exergy by energy carrier

The following figure is a recreation of Figure 7
of @Brockway:2024ab.

```{r fig7, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross useful energy and exergy consumption and shares by energy product group. Non-energy uses are excluded. This is a recreation (using CL-PFU database ', clpfu_version, ') of Figure 7 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).'), echo=FALSE}

# Build the data frame.
grouped_useful_data <- grouped_final_useful_data |> 
  dplyr::filter(
    IncludesNEU == FALSE,
    LastStage == "Useful") |> 
  dplyr::mutate(
    YUEIOU = matsbyname::rowsums_byname(YUEIOU)
  ) |> 
  tidyr::pivot_longer(cols = "YUEIOU", 
                      names_to = "matnames", 
                      values_to = "matvals") |>
  
  # This code is here only because I forgot to add
  # some heat categories to the eu_product aggregation map.
  # I have now added those aggregations, 
  # so the code below will not be needed in the future.
  # ---MKH, 10 July 2026
  dplyr::mutate(
    matvals = matsbyname::aggregate_byname(
      matvals,
      aggregation_map = list(list(
        HTH = c("HTH", "HTH.1200.C", "HTH.1800.C"), 
        MTH = c("MTH", "MTH.120.C", "MTH.130.C", "MTH.150.C", "MTH.160.C"), 
        LTH = c("LTH", "LTH.90.C")
      )), 
      margin = 1)
  ) |>
  
  matsindf::expand_to_tidy(drop = 0) |>
  # Adjust the names of useful energy carriers to be more readable
  dplyr::mutate(
    rownames = dplyr::case_when(
      rownames == "AnMech" ~ "Animal mechanical\nwork",
      rownames == "HPF" ~ "High-pressure\nfluids",
      rownames == "HTH" ~ "High-temperature\nheat",
      rownames == "HuMech" ~ "Human mechanical\nwork",
      rownames == "IP" ~ "Information\nprocessing",
      rownames == "L" ~ "Light",
      rownames == "LTC" ~ "Low-temperature\ncooling",
      rownames == "LTH" ~ "Low-temperature\nheat",
      rownames == "MechW" ~ "Mechanical work",
      rownames == "MTH" ~ "Medium-temperature\nheat",
      rownames == "NEU" ~ "Non-energy\nuse",
      rownames == "Prop" ~ "Propulsion",
      TRUE ~ .data[["rownames"]]),
    rownames = factor(rownames, 
                      levels = c("NEU", "Non-energy\nuse",
                                 "IP", "Information processing", "Information\nprocessing",
                                 "L", "Light",
                                 "HTH", "High-temperature heat", "High-temperature\nheat",
                                 "MTH", "Medium-temperature heat", "Medium-temperature\nheat",
                                 "LTH", "Low-temperature heat", "Low-temperature\nheat",
                                 "LTC", "Low-temperature cooling", "Low-temperature\ncooling",
                                 "HPF", "High-pressure fluids", "High-pressure\nfluids",
                                 "Prop", "Propulsion",
                                 "MechW", "Mechanical work",
                                 "HuMech", "Human mechanical work", "Human mechanical\nwork",
                                 "AnMech", "Animal mechanical work", "Animal mechanical\nwork"))
  )

# Make the graph
useful_ej <- grouped_useful_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 250, by = 50)) +
  ggplot2::scale_fill_manual(values = MKHthemes::useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme() + 
  ggplot2::theme(legend.text = element_text(size = legend_text_size),
                 axis.text.x = element_blank(),
                 axis.ticks.x = element_blank())


useful_fractions <- grouped_useful_data |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) +
  ggplot2::geom_area(position = ggplot2::position_fill()) +
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Shares [%]",
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  ggplot2::scale_fill_manual(values = MKHthemes::useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size),
                 strip.text.x = ggplot2::element_blank())

ggpubr::ggarrange(useful_ej,
                  useful_fractions,
                  nrow = 2,
                  common.legend = TRUE,
                  legend = "right",
                  align = "v")
```


```{r fig10, fig.align='center', fig.retina=2, fig.height = 5, fig.cap = paste0('World gross and net aggregate efficiencies including both IEA and muscle work (MW) energy data. Non-energy uses are excluded. $\\eta_{pf}$ is primary-to-final efficiency; $\\eta_{fu}$ is final-to-useful efficiency; $\\eta_{pu}$ is primary-to-useful efficiency. This is a recreation (using CL-PFU database ', clpfu_version, ') of Figure 10 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).'), echo=FALSE}
world_eta |> 
  dplyr::filter(LastStage == "Useful", IncludesNEU == FALSE, ProductAggregation == "Specified", IndustryAggregation == "Specified", Year >= 1971) |> 
  dplyr::select(-EXp, -EXf, -EXu) |> 
  tidyr::pivot_longer(cols = c(etapf, etafu, etapu), 
                      names_to = "VarName", 
                      values_to = "Value") |> 
  dplyr::mutate(
    Dataset = dplyr::case_when(
      Dataset == "CL-PFU IEA+MW" ~ "IEA + MW", 
      Dataset == "CL-PFU IEA" ~ "IEA", 
      Dataset == "CL-PFU MW" ~ "MW",
      TRUE ~ NA_character_
    ), 
    Dataset = factor(Dataset, levels = c("IEA", "MW", "IEA + MW")), 
    VarName = dplyr::case_when(
      VarName == "etapf" ~ "eta[pf]",
      VarName == "etafu" ~ "eta[fu]",
      VarName == "etapu" ~ "eta[pu]",
      TRUE ~ NA_character_
    ), 
    VarName = factor(VarName, levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = Value, 
                                         linetype = GrossNet, 
                                         colour = VarName)) +
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Dataset),
                      cols = ggplot2::vars(EnergyType),
                      scales = "free") + 
  ggplot2::scale_x_continuous(name = NULL,
                              breaks = c(1980, 2000, 2020)) + 
  ggplot2::scale_y_continuous(name = expression("Efficiency ("*eta*") [%]"),
                              limits = c(0, 1),
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  viridis::scale_color_viridis(name = NULL, 
                               direction = -1, 
                               discrete = TRUE, 
                               end = 0.85,
                               labels = scales::label_parse(),
                               option = "inferno") +
  ggplot2::scale_linetype_manual(name = NULL,
                                 values = c(Gross = "solid", 
                                            Net = "dashed")) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size))
```


# References
