---
title: "CL-PFU database pipeline diagnostic report"
output:
  bookdown::pdf_document2:
    toc: false
    citation_package: natbib
author: Matthew Kuperus Heun
date: "`r Sys.Date()`"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(ggpubr)
library(Matrix)
library(matsindf)
library(MKHthemes)
library(tidyr)

legend_text_size <- 8
```


# Introduction

This report provides diagnostics for the CL-PFU database
pipeline.
The purpose of this report is to compare key results 
from one version of the database to another.


# Preliminaries

We first set the version of the database
for which diagnostics are desired.

```{r version} 
clpfu_version <- "v3.0a2"
```

Set data file paths.

```{r data-file-paths}
world_data_file <- file.path("world_mats.rds")
world_eta_file <- file.path("world_eta.rds")
sector_agg_eta_file <- file.path("sector_agg_eta.rds")
```

We can download data using this chunk of code.
To save time, 
this chunk is not normally evaluated when building the report
(`eval=FALSE`).
Run this chunk as needed.

```{r download-data, eval=FALSE}
# We first make a connection to the database.
conn <- PFUPipelineTools::get_scratchmdb_conn()
on.exit(DBI::dbDisconnect(conn))

# Next, download desired data and save locally for later use. 
# We do this once to save time.
world_mats <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)

# Save to a file
world_mats |>
  saveRDS(file = world_data_file)

world_eta <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "AggEtaPFU", 
  Country == "World", 
  conn = conn, 
  collect = TRUE
)

world_eta |> 
  saveRDS(file = world_eta_file)

sector_agg_eta <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "SectorAggEtaFU", 
  Country == "World", 
  conn = conn, 
  collect = TRUE
)

sector_agg_eta |> 
  saveRDS(file = sector_agg_eta_file)

# Disconnect from the database management system.
DBI::dbDisconnect(conn)
```

Load data from the pre-saved file.

```{r load, echo=TRUE}
world_mats <- readRDS(file = world_data_file) |> 
  dplyr::mutate(
      EnergyType = dplyr::case_when(
      EnergyType == "E" ~ "Energy", 
      EnergyType == "X" ~ "Exergy", 
      TRUE ~ NA_character_
    ), 
    Dataset = dplyr::case_when(
      Dataset == "CL-PFU IEA+MW" ~ "IEA + MW", 
      Dataset == "CL-PFU IEA" ~ "IEA", 
      Dataset == "CL-PFU MW" ~ "MW",
      TRUE ~ NA_character_
    ), 
    Dataset = factor(Dataset, levels = c("IEA", "MW", "IEA + MW")), 
  )

world_eta <- readRDS(file = world_eta_file) |> 
  dplyr::mutate(
      EnergyType = dplyr::case_when(
      EnergyType == "E" ~ "Energy", 
      EnergyType == "X" ~ "Exergy", 
      TRUE ~ NA_character_
      ), 
      Dataset = dplyr::case_when(
        Dataset == "CL-PFU IEA+MW" ~ "IEA + MW", 
        Dataset == "CL-PFU IEA" ~ "IEA", 
        Dataset == "CL-PFU MW" ~ "MW",
        TRUE ~ NA_character_
      ), 
      Dataset = factor(Dataset, levels = c("IEA", "MW", "IEA + MW")), 
  )

sector_agg_eta <- readRDS(file = sector_agg_eta_file) |> 
  dplyr::mutate(
    EnergyType = dplyr::case_when(
      EnergyType == "E" ~ "Energy", 
      EnergyType == "X" ~ "Exergy", 
      TRUE ~ NA_character_
    ), 
    Dataset = dplyr::case_when(
      Dataset == "CL-PFU IEA+MW" ~ "IEA + MW", 
      Dataset == "CL-PFU IEA" ~ "IEA", 
      Dataset == "CL-PFU MW" ~ "MW",
      TRUE ~ NA_character_
    ), 
    Dataset = factor(Dataset, levels = c("IEA", "MW", "IEA + MW")), 
  )
```

At this point, we are ready to make diagnostic graphs and tables.


# Primary, final, and useful aggregations

## Global primary and final energy and exergy by sector

Figure \@ref(fig:fig5) is a recreation of Figure 5
of @Brockway:2024ab.

```{r fig5, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross primary and final energy and exergy consumption by energy product group. Non-energy uses are included. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 5 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).'), echo='FALSE'}
# Define sectors to exclude from the data
# to match sectors excluded from Figures 5, 6, and 7.
sectors_to_exclude <- RCLabels::make_or_pattern(
  c("Exports", 
    "Exports to World aviation bunkers", 
    "Exports to World marine bunkers", 
    "Losses", 
    "Statistical differences", 
    "Stock changes", 
    "Imports"))

RYUEIOU <- world_mats |> 
  # Filter to the same parameters as the paper's graphs
  # for following figures.
  dplyr::filter(Dataset == "IEA + MW",
                Year >= 1971,
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType, LastStage) |> 
  dplyr::mutate(
    R = R |> 
      # For world level aggregates, we should not include imports in R
      matsbyname::select_rows_byname(remove_pattern = sectors_to_exclude) |>
      # Convert R from TJ to EJ
      matsbyname::hadamardproduct_byname(1e-6),
    # For gross energy, we need to include both Y and U_EIOU matrices
    YUEIOU = Y |> 
      matsbyname::sum_byname(U_EIOU) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)  |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU") |>
      # Convert YUEIOU from TJ to EJ
      matsbyname::hadamardproduct_byname(1e-6)
  )

RYUEIOU |> 
  dplyr::filter(LastStage == "Final",
                IncludesNEU == TRUE) |> 
  dplyr::select(-U, -U_EIOU, -U_feed, -r_EIOU, -V, -Y, -S_units) |> 
  dplyr::mutate(
    # Put R and Y in the same orientation to make future
    # manipulations easier.
    R = matsbyname::transpose_byname(R)
  ) |> 
  tidyr::pivot_longer(cols = c("R", "YUEIOU"), 
                      names_to = "matnames",
                      values_to = "matvals") |>
  dplyr::mutate(
    matvals = matsbyname::rowsums_byname(matvals),
    Stage = dplyr::case_when(
      matnames == "R" ~ "Primary", 
      matnames == "YUEIOU" ~ "Final", 
      TRUE ~ NA_character_
    ), 
    Stage = factor(Stage, levels = c("Primary", "Final"))
  ) |>
  matsindf::expand_to_tidy() |> 
  dplyr::mutate(
    rownames = dplyr::case_when(
      rownames == "Oil & oil products" ~ "Oil & oil\nproducts", 
      rownames == "Coal & coal products" ~ "Coal & coal\nproducts", 
      TRUE ~ rownames
    ),
    rownames = factor(rownames, 
                      levels = c("Waste", "Heat", "Electricity", 
                                 "Biofuels", "Natural gas",
                                 "Oil & oil\nproducts", 
                                 "Coal & coal\nproducts", 
                                 "Biomass", "Food", "Feed"))
    
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Stage), 
                      cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(name = NULL,
                              breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Energy and exergy [EJ/year]") +
  ggplot2::scale_fill_manual(name = NULL, 
                             values = MKHthemes::primary_and_final_carrier_colours) +
  MKHthemes::xy_theme()
```


## Global final and useful energy and exergy by sector

Figure \@ref(fig:fig6) is a recreation of Figure 6
of @Brockway:2024ab.

```{r fig6, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross final and useful energy and exergy consumption by sector. Non-energy uses are included at the final energy stage. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 6 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).'), echo=FALSE}
grouped_final_useful_data <- RYUEIOU |> 
  dplyr::select(-R, -U, -U_feed, -U_EIOU, -r_EIOU, -V, -Y, -S_units)

grouped_final_useful_data |> 
  dplyr::mutate(
    YUEIOU = matsbyname::colsums_byname(YUEIOU)
  ) |> 
  tidyr::pivot_longer(cols = "YUEIOU", 
                      names_to = "matnames", 
                      values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  dplyr::filter(IncludesNEU == TRUE) |> 
  dplyr::mutate(
    colnames = dplyr::case_when(
      colnames == "Agriculture, forestry, and fishing" ~ "Agriculture, forestry,\nand fishing",
      colnames == "Commercial and public services" ~ "Commercial and\npublic services",
      colnames == "Final consumption not elsewhere specified" ~ "Final consumption not\nelsewhere specified",
      colnames == "Energy industry own use" ~ "Energy industry\nown use",
      TRUE ~ .data[["colnames"]]
    ), 
    colnames = factor(colnames,
                      levels = c("Energy industry\nown use",
                                 "Non-energy use",
                                 "Final consumption not\nelsewhere specified",
                                 "Transport",
                                 "Commercial and\npublic services",
                                 "Industry",
                                 "Residential",
                                 "Agriculture, forestry,\nand fishing"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = colnames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(rows = ggplot2::vars(LastStage), 
                      cols = ggplot2::vars(EnergyType)) + 
  ggplot2::labs(x = NULL, 
                fill = NULL) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Energy and exergy [EJ/year]") +
  ggplot2::scale_fill_manual(values = MKHthemes::sector_colours) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size))
```

## Global useful energy and exergy by energy carrier

Figure \@ref(fig:fig7) is a recreation of Figure 7
of @Brockway:2024ab.

```{r fig7, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross useful energy and exergy consumption and shares by energy product group. Non-energy uses are excluded. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 7 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).'), echo=FALSE}
# Build the data frame.
grouped_useful_data <- grouped_final_useful_data |> 
  dplyr::filter(
    IncludesNEU == FALSE,
    LastStage == "Useful") |> 
  dplyr::mutate(
    YUEIOU = matsbyname::rowsums_byname(YUEIOU)
  ) |> 
  tidyr::pivot_longer(cols = "YUEIOU", 
                      names_to = "matnames", 
                      values_to = "matvals") |>
  
  # This code is here only because I forgot to add
  # some heat categories to the eu_product aggregation map.
  # I have now added those aggregations, 
  # so the code below will not be needed in the future.
  # ---MKH, 10 July 2026
  dplyr::mutate(
    matvals = matsbyname::aggregate_byname(
      matvals,
      aggregation_map = list(list(
        HTH = c("HTH", "HTH.1200.C", "HTH.1800.C"), 
        MTH = c("MTH", "MTH.120.C", "MTH.130.C", "MTH.150.C", "MTH.160.C"), 
        LTH = c("LTH", "LTH.90.C")
      )), 
      margin = 1)
  ) |>
  
  matsindf::expand_to_tidy(drop = 0) |>
  # Adjust the names of useful energy carriers to be more readable
  dplyr::mutate(
    rownames = dplyr::case_when(
      rownames == "AnMech" ~ "Animal mechanical\nwork",
      rownames == "HPF" ~ "High-pressure\nfluids",
      rownames == "HTH" ~ "High-temperature\nheat",
      rownames == "HuMech" ~ "Human mechanical\nwork",
      rownames == "IP" ~ "Information\nprocessing",
      rownames == "L" ~ "Light",
      rownames == "LTC" ~ "Low-temperature\ncooling",
      rownames == "LTH" ~ "Low-temperature\nheat",
      rownames == "MechW" ~ "Mechanical work",
      rownames == "MTH" ~ "Medium-temperature\nheat",
      rownames == "NEU" ~ "Non-energy\nuse",
      rownames == "Prop" ~ "Propulsion",
      TRUE ~ .data[["rownames"]]),
    rownames = factor(rownames, 
                      levels = c("NEU", "Non-energy\nuse",
                                 "IP", "Information processing", "Information\nprocessing",
                                 "L", "Light",
                                 "HTH", "High-temperature heat", "High-temperature\nheat",
                                 "MTH", "Medium-temperature heat", "Medium-temperature\nheat",
                                 "LTH", "Low-temperature heat", "Low-temperature\nheat",
                                 "LTC", "Low-temperature cooling", "Low-temperature\ncooling",
                                 "HPF", "High-pressure fluids", "High-pressure\nfluids",
                                 "Prop", "Propulsion",
                                 "MechW", "Mechanical work",
                                 "HuMech", "Human mechanical work", "Human mechanical\nwork",
                                 "AnMech", "Animal mechanical work", "Animal mechanical\nwork"))
  )

# Make the graph
useful_ej <- grouped_useful_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 250, by = 50)) +
  ggplot2::scale_fill_manual(values = MKHthemes::useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme() + 
  ggplot2::theme(legend.text = element_text(size = legend_text_size),
                 axis.text.x = element_blank(),
                 axis.ticks.x = element_blank())


useful_fractions <- grouped_useful_data |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) +
  ggplot2::geom_area(position = ggplot2::position_fill()) +
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Shares [%]",
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  ggplot2::scale_fill_manual(values = MKHthemes::useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size),
                 strip.text.x = ggplot2::element_blank())

ggpubr::ggarrange(useful_ej,
                  useful_fractions,
                  nrow = 2,
                  common.legend = TRUE,
                  legend = "right",
                  align = "v")
```


Figure \@ref(fig:fig8) is a recreation of Figure 8
of @Brockway:2024ab.

```{r fig8, fig.height=5, fig.align='center', fig.retina=2, echo=FALSE, fig.cap=paste0('Gross and net world aggregate energy and exergy consumption including both IEA and muscle work (MW) energy data. Note energy and exergy are identical for muscle work, and the scale is different for the muscle work (MW) graphs. Non-energy uses are excluded. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 8 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).')}
world_eta |> 
  dplyr::select(-etapf, -etafu, -etapu) |>
  dplyr::filter(Year >= 1971, 
                IncludesNEU == FALSE) |> 
  tidyr::pivot_longer(cols = c("EXp", "EXf", "EXu"), 
                      names_to = "varname", 
                      values_to = "value") |> 
  dplyr::mutate(
    value = value / 1e6, # Convert from TJ to EJ
    varname = dplyr::case_when(
      varname == "EXp" ~ "Primary", 
      varname == "EXf" ~ "Final", 
      varname == "EXu" ~ "Useful", 
      TRUE ~ NA_character_
    ),
    varname = factor(varname, levels = c("Primary", "Final", "Useful"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = value,
                                         colour = varname, 
                                         linetype = GrossNet)) + 
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Dataset), 
                      cols = ggplot2::vars(EnergyType), 
                      scales = "free") + 
  ggplot2::scale_x_continuous(name = NULL, breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Energy and exergy [EJ/year]") +
  ggplot2::scale_linetype(name = NULL) +
  viridis::scale_color_viridis(name = NULL, 
                               direction = -1, 
                               discrete = TRUE, 
                               end = 0.7) +
  MKHthemes::xy_theme()
```

## World efficiencies

Figure \@ref(fig:fig10) is a recreation of Figure 10
of @Brockway:2024ab.

```{r fig10, fig.align='center', fig.retina=2, fig.height=5, echo=FALSE, fig.cap=paste0('World gross and net aggregate efficiencies including both IEA and muscle work (MW) energy data. Non-energy uses are excluded. $\\eta_{pf}$ is primary-to-final efficiency; $\\eta_{fu}$ is final-to-useful efficiency; $\\eta_{pu}$ is primary-to-useful efficiency. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 10 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).')}
world_eta_for_fig_10 <- world_eta |> 
  dplyr::filter(LastStage == "Useful", 
                IncludesNEU == FALSE,
                ProductAggregation == "Specified",
                IndustryAggregation == "Specified", Year >= 1971) |> 
  dplyr::select(-EXp, -EXf, -EXu) |> 
  tidyr::pivot_longer(cols = c(etapf, etafu, etapu), 
                      names_to = "VarName", 
                      values_to = "Value") |> 
  dplyr::mutate(
    VarName = dplyr::case_when(
      VarName == "etapf" ~ "eta[pf]",
      VarName == "etafu" ~ "eta[fu]",
      VarName == "etapu" ~ "eta[pu]",
      TRUE ~ NA_character_
    ), 
    VarName = factor(VarName, levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  )

world_eta_for_fig_10 |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = Value, 
                                         linetype = GrossNet, 
                                         colour = VarName)) +
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Dataset),
                      cols = ggplot2::vars(EnergyType),
                      scales = "free") + 
  ggplot2::scale_x_continuous(name = NULL,
                              breaks = c(1980, 2000, 2020)) + 
  ggplot2::scale_y_continuous(name = expression("Efficiency ("*eta*") [%]"),
                              limits = c(0, 1),
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  viridis::scale_color_viridis(name = NULL, 
                               direction = -1, 
                               discrete = TRUE, 
                               end = 0.85,
                               labels = scales::label_parse(),
                               option = "inferno") +
  ggplot2::scale_linetype_manual(name = NULL,
                                 values = c(Gross = "solid", 
                                            Net = "dashed")) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size))
```


Figure \@ref(fig:fig11) is a recreation of Figure 11
of @Brockway:2024ab.

```{r fig11, fig.align='center', fig.retina=2, fig.height=5, echo=FALSE, fig.cap=paste0('World gross FU efficiency ($\\eta_{fu}$) by sector. Non-energy uses are excluded. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 10 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).')}
dplyr::bind_rows(
  # Get the world average efficiencies
  world_eta_for_fig_10 |> 
    dplyr::filter(VarName == "eta[fu]", 
                  GrossNet == "Gross", 
                  IncludesNEU == FALSE) |> 
    dplyr::select(-VarName) |> 
    dplyr::rename(etafu = "Value") |> 
    dplyr::mutate(
      Sector = "World",
      LastStage = NULL
    ), 
  sector_agg_eta |> 
    dplyr::filter(Year >= 1971, 
                  ProductAggregation == "Grouped", 
                  IndustryAggregation == "Grouped", 
                  GrossNet == "Gross", 
                  IncludesNEU == FALSE) |> 
    dplyr::select(-Final, -Useful)
) |>
  dplyr::mutate(
    Dataset = factor(Dataset, 
                     levels = c("IEA", "MW", "IEA + MW")), 
    Sector = dplyr::case_when(
      Sector == "Agriculture, forestry, and fishing" ~ "Agriculture, forestry,\nand fishing",
      Sector == "Commercial and public services" ~ "Commercial and\npublic services",
      Sector == "Final consumption not elsewhere specified" ~ "Final consumption not\nelsewhere specified",
      Sector == "Energy industry own use" ~ "Energy industry\nown use",
      TRUE ~ .data[["Sector"]]
    ),
    Sector = factor(Sector,
                    levels = c("Energy industry\nown use",
                               "Non-energy use",
                               "Final consumption not\nelsewhere specified",
                               "Transport",
                               "Commercial and\npublic services",
                               "Industry",
                               "Residential",
                               "Agriculture, forestry,\nand fishing",
                               "World"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = etafu, 
                                         colour = Sector)) + 
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Dataset), 
                      cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(name = NULL, 
                              breaks = c(1980, 2000, 2020)) + 
  ggplot2::scale_y_continuous(name = expression("Final-to-useful efficiency ("*eta[fu]*") [%]"),
                              limits = c(0, NA),
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  ggplot2::scale_colour_manual(name = NULL, 
                               values = MKHthemes::sector_colours) +
  
  
  MKHthemes::xy_theme()
```



Table \@ref(tab:tab4) is a recreation of Table 4
of @Brockway:2024ab.


```{r tab4, echo=FALSE, tab.cap = paste0('Gross world efficiencies including both IEA and muscle work data. Non-energy uses are excluded. This is a recreation (using CL-PFU database ', clpfu_version, ') of Table 4 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).')}
world_eta |> 
  dplyr::filter(Dataset == "IEA + MW", 
                Year %in% c(1971, 2020), 
                LastStage == "Useful",
                IncludesNEU == FALSE, 
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified", 
                GrossNet == "Gross") |>
  dplyr::select(-EXp, -EXf, -EXu) |>
  tidyr::pivot_longer(cols = c("etapf", "etafu", "etapu"),
                      names_to = "variable",
                      values_to = "value") |>
  dplyr::mutate(
    coltitle = paste0(EnergyType, "-", Year), 
    EnergyType = NULL, 
    Year = NULL, 
    variable = dplyr::case_when(
      variable == "etapf" ~ "eta[pf]", 
      variable == "etafu" ~ "eta[fu]", 
      variable == "etapu" ~ "eta[pu]", 
      TRUE ~ NA_character_
    )
  ) |> 
  tidyr::pivot_wider(names_from = "coltitle", values_from = "value") |>
  dplyr::select("variable", 
                "Energy-1971", "Energy-2020", 
                "Exergy-1971", "Exergy-2020") |>  
  dplyr::mutate(
    variable = factor(variable, 
                      levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  ) |> 
  dplyr::mutate(
    # Convert to %
    dplyr::across(where(is.numeric),
                  ~ sprintf("%.2f %%", .x * 100)
                  )
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE)
```

Table \@ref(tab:tab5) is a recreation of Table 5
of @Brockway:2024ab.

```{r tab5, tab.cap=paste0('Gross world primary, final, and useful energy and exergy growth metrics. Both IEA and muscle work data are included. Non-energy uses are excluded. This is a recreation (using CL-PFU database ', clpfu_version, ') of Table 5 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.3).'), echo=FALSE}
world_eta |> 
  dplyr::filter(Dataset == "IEA + MW", 
                Year %in% c(1971, 2020), 
                LastStage == "Useful",
                IncludesNEU == FALSE, 
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified", 
                GrossNet == "Gross") |>
  dplyr::select(-etapf, -etafu, -etapu) |> 
  tidyr::pivot_longer(cols = c("EXp", "EXf", "EXu"),
                      names_to = "variable",
                      values_to = "value") |>
  dplyr::mutate(
    coltitle = paste0(EnergyType, "-", Year), 
    EnergyType = NULL, 
    Year = NULL, 
  ) |>
  tidyr::pivot_wider(names_from = "coltitle", values_from = "value") |>
  dplyr::select("variable", 
                "Energy-1971", "Energy-2020", 
                "Exergy-1971", "Exergy-2020") |> 
  dplyr::mutate(
    variable = dplyr::case_when(
      variable == "EXp" ~ "Primary", 
      variable == "EXf" ~ "Final", 
      variable == "EXu" ~ "Useful", 
      TRUE ~ NA_character_
    ), 
    `Energy 2020/1971` = sprintf(fmt = "%.2f x", 
                                 `Energy-2020` / `Energy-1971`), 
    `Exergy 2020/1971` = sprintf(fmt = "%.2f x", 
                                 `Exergy-2020` / `Exergy-1971`),
    `r Energy 2020/1971` = sprintf(fmt = "%.2f %%/yr", 
                                    ((`Energy-2020`/`Energy-1971`) ^ 
                                       (1/(2020-1971)) - 1) * 100),
    `r Exergy 2020/1971` = sprintf(fmt = "%.2f %%/yr",
                                    ((`Exergy-2020`/`Exergy-1971`) ^
                                       (1/(2020-1971)) - 1) * 100), 
    `Energy-1971` = NULL, 
    `Energy-2020` = NULL, 
    `Exergy-1971` = NULL, 
    `Exergy-2020` = NULL
  ) |> 
  knitr::kable(format = "latex", booktabs = TRUE)
```


# References
