---
title: "CL-PFU database pipeline diagnostic report"
output:
  bookdown::pdf_document2:
    toc: false
    citation_package: natbib
author: Matthew Kuperus Heun
date: "`r Sys.Date()`"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggh4x)
library(ggplot2)
library(ggpubr)
library(Matrix)
library(matsindf)
library(MKHthemes)
library(scales)
library(tidyr)

legend_text_size <- 8
```


# Introduction

This report provides diagnostics for the CL-PFU database
pipeline.
The purpose of this report is to compare key results 
from one version of the database to another.


# Preliminaries

We first set the version of the database
for which diagnostics are desired.

```{r version} 
clpfu_version <- "v3.0b1"
```

Set data file paths.

```{r data-file-paths}
world_data_file <- file.path("world_mats.rds")
world_continent_eta_file <- file.path("world_continent_eta.rds")
sector_agg_eta_file <- file.path("sector_agg_eta.rds")
completed_allocation_tables_file <- file.path("completed_allocation_tables.rds")
completed_efficiency_tables_file <- file.path("completed_efficiency_tables.rds")
```

We can download data using this chunk of code.
To save time, 
this chunk is not normally evaluated when building the report
(`eval=FALSE`).
Run this chunk as needed.

```{r download-data, eval=FALSE}
# We first make a connection to the database.
conn <- PFUPipelineTools::get_scratchmdb_conn()
on.exit(DBI::dbDisconnect(conn))

# Next, download desired data and save locally for later use. 
# We do this once to save time.
world_mats <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)

# Save to a file
world_mats |>
  saveRDS(file = world_data_file)

world_continent_eta <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "AggEtaPFU", 
  # Include GRC for double-checking the previous FAO error
  Country %in% c("GRC", "Africa", "Asia_", "Europe", "MidEast", "NoAmr", "Oceania", "SoCeAmr", "World"),
  conn = conn, 
  collect = TRUE
)

world_continent_eta |> 
  saveRDS(file = world_continent_eta_file)

sector_agg_eta <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "SectorAggEtaFU", 
  Country == "World", 
  conn = conn, 
  collect = TRUE
)

sector_agg_eta |> 
  saveRDS(file = sector_agg_eta_file)

completed_allocation_tables <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "CompletedAllocationTables", 
  conn = conn, 
  collect = TRUE
)

completed_allocation_tables |> 
  saveRDS(file = completed_allocation_tables_file)

completed_efficiency_tables <- PFUPipelineTools::pl_filter_collect(
  version_string = clpfu_version, 
  db_table_name = "CompletedEfficiencyTables", 
  conn = conn, 
  collect = TRUE
)

completed_efficiency_tables |> 
  saveRDS(file = completed_efficiency_tables_file)

# Disconnect from the database management system.
DBI::dbDisconnect(conn)
```

Load data from the pre-saved files.

```{r load, echo=TRUE}
world_mats <- readRDS(file = world_data_file) |> 
  dplyr::mutate(
      EnergyType = dplyr::case_when(
      EnergyType == "E" ~ "Energy", 
      EnergyType == "X" ~ "Exergy", 
      TRUE ~ NA_character_
    ), 
    Dataset = dplyr::case_when(
      Dataset == "CL-PFU IEA+MW" ~ "IEA + MW", 
      Dataset == "CL-PFU IEA" ~ "IEA", 
      Dataset == "CL-PFU MW" ~ "MW",
      TRUE ~ NA_character_
    ), 
    Dataset = factor(Dataset, levels = c("IEA", "MW", "IEA + MW")), 
  )

world_continent_eta <- readRDS(file = world_continent_eta_file) |> 
  dplyr::mutate(
      EnergyType = dplyr::case_when(
      EnergyType == "E" ~ "Energy", 
      EnergyType == "X" ~ "Exergy", 
      TRUE ~ NA_character_
      ), 
      Dataset = dplyr::case_when(
        Dataset == "CL-PFU IEA+MW" ~ "IEA + MW", 
        Dataset == "CL-PFU IEA" ~ "IEA", 
        Dataset == "CL-PFU MW" ~ "MW",
        TRUE ~ NA_character_
      ), 
      Dataset = factor(Dataset, levels = c("IEA", "MW", "IEA + MW")), 
  )

sector_agg_eta <- readRDS(file = sector_agg_eta_file) |> 
  dplyr::mutate(
    EnergyType = dplyr::case_when(
      EnergyType == "E" ~ "Energy", 
      EnergyType == "X" ~ "Exergy", 
      TRUE ~ NA_character_
    ), 
    Dataset = dplyr::case_when(
      Dataset == "CL-PFU IEA+MW" ~ "IEA + MW", 
      Dataset == "CL-PFU IEA" ~ "IEA", 
      Dataset == "CL-PFU MW" ~ "MW",
      TRUE ~ NA_character_
    ), 
    Dataset = factor(Dataset, levels = c("IEA", "MW", "IEA + MW")), 
  )

completed_allocation_tables <- readRDS(file = completed_allocation_tables_file)

completed_efficiency_tables <- readRDS(file = completed_efficiency_tables_file)
```

At this point, we are ready to make diagnostic graphs and tables.


# Primary, final, and useful aggregations

## Global primary and final energy and exergy by sector

Figure \@ref(fig:fig5) is a recreation of Figure 5
of @Brockway:2024ab.

```{r fig5, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross primary and final energy and exergy consumption by energy product group. Non-energy uses are included. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 5 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).'), echo='FALSE'}
# Define sectors to exclude from the data
# to match sectors excluded from Figures 5, 6, and 7.
sectors_to_exclude <- RCLabels::make_or_pattern(
  c("Exports", 
    "Exports to World aviation bunkers", 
    "Exports to World marine bunkers", 
    "Losses", 
    "Statistical differences", 
    "Stock changes", 
    "Imports"))

RYUEIOU <- world_mats |> 
  # Filter to the same parameters as the paper's graphs
  # for following figures.
  dplyr::filter(Dataset == "IEA + MW",
                Year >= 1971,
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType, LastStage) |> 
  dplyr::mutate(
    R = R |> 
      # For world level aggregates, we should not include imports in R
      matsbyname::select_rows_byname(remove_pattern = sectors_to_exclude) |>
      # Convert R from TJ to EJ
      matsbyname::hadamardproduct_byname(1e-6),
    # For gross energy, we need to include both Y and U_EIOU matrices
    YUEIOU = Y |> 
      matsbyname::sum_byname(U_EIOU) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)  |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU") |>
      # Convert YUEIOU from TJ to EJ
      matsbyname::hadamardproduct_byname(1e-6)
  )

RYUEIOU |> 
  dplyr::filter(LastStage == "Final",
                IncludesNEU == TRUE) |> 
  dplyr::select(-U, -U_EIOU, -U_feed, -r_EIOU, -V, -Y, -S_units) |> 
  dplyr::mutate(
    # Put R and Y in the same orientation to make future
    # manipulations easier.
    R = matsbyname::transpose_byname(R)
  ) |> 
  tidyr::pivot_longer(cols = c("R", "YUEIOU"), 
                      names_to = "matnames",
                      values_to = "matvals") |>
  dplyr::mutate(
    matvals = matsbyname::rowsums_byname(matvals),
    Stage = dplyr::case_when(
      matnames == "R" ~ "Primary", 
      matnames == "YUEIOU" ~ "Final", 
      TRUE ~ NA_character_
    ), 
    Stage = factor(Stage, levels = c("Primary", "Final"))
  ) |>
  matsindf::expand_to_tidy() |> 
  dplyr::mutate(
    rownames = dplyr::case_when(
      rownames == "Oil & oil products" ~ "Oil & oil\nproducts", 
      rownames == "Coal & coal products" ~ "Coal & coal\nproducts", 
      TRUE ~ rownames
    ),
    rownames = factor(rownames, 
                      levels = c("Waste", "Heat", "Electricity", 
                                 "Biofuels", "Natural gas",
                                 "Oil & oil\nproducts", 
                                 "Coal & coal\nproducts", 
                                 "Biomass", "Food", "Feed"))
    
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Stage), 
                      cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(name = NULL,
                              breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Energy and exergy [EJ/year]") +
  ggplot2::scale_fill_manual(name = NULL, 
                             values = MKHthemes::primary_and_final_carrier_colours) +
  MKHthemes::xy_theme()
```


## Global final and useful energy and exergy by sector

Figure \@ref(fig:fig6) is a recreation of Figure 6
of @Brockway:2024ab.

```{r fig6, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross final and useful energy and exergy consumption by sector. Non-energy uses are included at the final energy stage. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 6 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).'), echo=FALSE}
grouped_final_useful_data <- RYUEIOU |> 
  dplyr::select(-R, -U, -U_feed, -U_EIOU, -r_EIOU, -V, -Y, -S_units)

grouped_final_useful_data |> 
  dplyr::mutate(
    YUEIOU = matsbyname::colsums_byname(YUEIOU)
  ) |> 
  tidyr::pivot_longer(cols = "YUEIOU", 
                      names_to = "matnames", 
                      values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  dplyr::filter(IncludesNEU == TRUE) |> 
  dplyr::mutate(
    colnames = dplyr::case_when(
      colnames == "Agriculture, forestry, and fishing" ~ "Agriculture, forestry,\nand fishing",
      colnames == "Commercial and public services" ~ "Commercial and\npublic services",
      colnames == "Final consumption not elsewhere specified" ~ "Final consumption not\nelsewhere specified",
      colnames == "Energy industry own use" ~ "Energy industry\nown use",
      TRUE ~ .data[["colnames"]]
    ), 
    colnames = factor(colnames,
                      levels = c("Energy industry\nown use",
                                 "Non-energy use",
                                 "Final consumption not\nelsewhere specified",
                                 "Transport",
                                 "Commercial and\npublic services",
                                 "Industry",
                                 "Residential",
                                 "Agriculture, forestry,\nand fishing"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = colnames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(rows = ggplot2::vars(LastStage), 
                      cols = ggplot2::vars(EnergyType)) + 
  ggplot2::labs(x = NULL, 
                fill = NULL) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Energy and exergy [EJ/year]") +
  ggplot2::scale_fill_manual(values = MKHthemes::sector_colours) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size))
```

## Global useful energy and exergy by energy carrier

Figure \@ref(fig:fig7) is a recreation of Figure 7
of @Brockway:2024ab.

```{r fig7, fig.height=5, fig.align='center', fig.retina=2, fig.cap=paste0('World gross useful energy and exergy consumption and shares by energy product group. Non-energy uses are excluded. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 7 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).'), echo=FALSE}
# Build the data frame.
grouped_useful_data <- grouped_final_useful_data |> 
  dplyr::filter(
    IncludesNEU == FALSE,
    LastStage == "Useful") |> 
  dplyr::mutate(
    YUEIOU = matsbyname::rowsums_byname(YUEIOU)
  ) |> 
  tidyr::pivot_longer(cols = "YUEIOU", 
                      names_to = "matnames", 
                      values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  # Adjust the names of useful energy carriers to be more readable
  dplyr::mutate(
    rownames = dplyr::case_when(
      rownames == "AnMech" ~ "Animal mechanical\nwork",
      rownames == "HPF" ~ "High-pressure\nfluids",
      rownames == "HTH" ~ "High-temperature\nheat",
      rownames == "HuMech" ~ "Human mechanical\nwork",
      rownames == "IP" ~ "Information\nprocessing",
      rownames == "L" ~ "Light",
      rownames == "LTC" ~ "Low-temperature\ncooling",
      rownames == "LTH" ~ "Low-temperature\nheat",
      rownames == "MechW" ~ "Mechanical work",
      rownames == "MTH" ~ "Medium-temperature\nheat",
      rownames == "NEU" ~ "Non-energy\nuse",
      rownames == "Prop" ~ "Propulsion",
      TRUE ~ .data[["rownames"]]),
    rownames = factor(rownames, 
                      levels = c("NEU", "Non-energy\nuse",
                                 "IP", "Information processing", "Information\nprocessing",
                                 "L", "Light",
                                 "HTH", "High-temperature heat", "High-temperature\nheat",
                                 "MTH", "Medium-temperature heat", "Medium-temperature\nheat",
                                 "LTH", "Low-temperature heat", "Low-temperature\nheat",
                                 "LTC", "Low-temperature cooling", "Low-temperature\ncooling",
                                 "HPF", "High-pressure fluids", "High-pressure\nfluids",
                                 "Prop", "Propulsion",
                                 "MechW", "Mechanical work",
                                 "HuMech", "Human mechanical work", "Human mechanical\nwork",
                                 "AnMech", "Animal mechanical work", "Animal mechanical\nwork"))
  )

# Make the graph
useful_ej <- grouped_useful_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 250, by = 50)) +
  ggplot2::scale_fill_manual(values = MKHthemes::useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme() + 
  ggplot2::theme(legend.text = element_text(size = legend_text_size),
                 axis.text.x = element_blank(),
                 axis.ticks.x = element_blank())


useful_fractions <- grouped_useful_data |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) +
  ggplot2::geom_area(position = ggplot2::position_fill()) +
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Shares [%]",
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  ggplot2::scale_fill_manual(values = MKHthemes::useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size),
                 strip.text.x = ggplot2::element_blank())

ggpubr::ggarrange(useful_ej,
                  useful_fractions,
                  nrow = 2,
                  common.legend = TRUE,
                  legend = "right",
                  align = "v")
```


Figure \@ref(fig:fig8) is a recreation of Figure 8
of @Brockway:2024ab.

```{r fig8, fig.height=5, fig.align='center', fig.retina=2, echo=FALSE, fig.cap=paste0('Gross and net world aggregate energy and exergy consumption including both IEA and muscle work (MW) energy data. Note energy and exergy are identical for muscle work, and the scale is different for the muscle work (MW) graphs. Non-energy uses are excluded. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 8 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).')}
fig8_data <- world_continent_eta |> 
  dplyr::select(-etapf, -etafu, -etapu) |>
  dplyr::filter(Country == "World", 
                Year >= 1971, 
                IncludesNEU == FALSE) |> 
  tidyr::pivot_longer(cols = c("EXp", "EXf", "EXu"), 
                      names_to = "varname", 
                      values_to = "value") |> 
  dplyr::mutate(
    value = value / 1e6, # Convert from TJ to EJ
    varname = dplyr::case_when(
      varname == "EXp" ~ "Primary", 
      varname == "EXf" ~ "Final", 
      varname == "EXu" ~ "Useful", 
      TRUE ~ NA_character_
    ),
    varname = factor(varname, levels = c("Primary", "Final", "Useful"))
  )

fig8_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = value,
                                         colour = varname, 
                                         linetype = GrossNet)) + 
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Dataset), 
                      cols = ggplot2::vars(EnergyType), 
                      scales = "free") + 
  ggplot2::scale_x_continuous(name = NULL, breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Energy and exergy [EJ/year]") +
  ggplot2::scale_linetype(name = NULL) +
  viridis::scale_color_viridis(name = NULL, 
                               direction = -1, 
                               discrete = TRUE, 
                               end = 0.7) +
  MKHthemes::xy_theme()
```

## World efficiencies

Figure \@ref(fig:fig10) is a recreation of Figure 10
of @Brockway:2024ab.

```{r fig10, fig.align='center', fig.retina=2, fig.height=5, echo=FALSE, fig.cap=paste0('World gross and net aggregate efficiencies including both IEA and muscle work (MW) energy data. Non-energy uses are excluded. $\\eta_{pf}$ is primary-to-final efficiency; $\\eta_{fu}$ is final-to-useful efficiency; $\\eta_{pu}$ is primary-to-useful efficiency. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 10 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).')}
world_eta_for_fig_10 <- world_continent_eta |> 
  dplyr::filter(Country == "World", 
                LastStage == "Useful", 
                IncludesNEU == FALSE,
                ProductAggregation == "Specified",
                IndustryAggregation == "Specified", Year >= 1971) |> 
  dplyr::select(-EXp, -EXf, -EXu) |> 
  tidyr::pivot_longer(cols = c(etapf, etafu, etapu), 
                      names_to = "VarName", 
                      values_to = "Value") |> 
  dplyr::mutate(
    VarName = dplyr::case_when(
      VarName == "etapf" ~ "eta[pf]",
      VarName == "etafu" ~ "eta[fu]",
      VarName == "etapu" ~ "eta[pu]",
      TRUE ~ NA_character_
    ), 
    VarName = factor(VarName, levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  )

world_eta_for_fig_10 |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = Value, 
                                         linetype = GrossNet, 
                                         colour = VarName)) +
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Dataset),
                      cols = ggplot2::vars(EnergyType),
                      scales = "free") + 
  ggplot2::scale_x_continuous(name = NULL,
                              breaks = c(1980, 2000, 2020)) + 
  ggplot2::scale_y_continuous(name = expression("Efficiency ("*eta*") [%]"),
                              limits = c(0, 1),
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  viridis::scale_color_viridis(name = NULL, 
                               direction = -1, 
                               discrete = TRUE, 
                               end = 0.85,
                               labels = scales::label_parse(),
                               option = "inferno") +
  ggplot2::scale_linetype_manual(name = NULL,
                                 values = c(Gross = "solid", 
                                            Net = "dashed")) +
  MKHthemes::xy_theme() +
  ggplot2::theme(legend.text = element_text(size = legend_text_size))
```


Figure \@ref(fig:fig11) is a recreation of Figure 11
of @Brockway:2024ab.

```{r fig11, fig.align='center', fig.retina=2, fig.height=5, echo=FALSE, fig.cap=paste0('World gross FU efficiency ($\\eta_{fu}$) by sector. Non-energy uses are excluded. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 11 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).')}
dplyr::bind_rows(
  # Get the world average efficiencies
  world_eta_for_fig_10 |> 
    dplyr::filter(VarName == "eta[fu]", 
                  GrossNet == "Gross", 
                  IncludesNEU == FALSE) |> 
    dplyr::select(-VarName) |> 
    dplyr::rename(etafu = "Value") |> 
    dplyr::mutate(
      Sector = "World",
      LastStage = NULL
    ), 
  sector_agg_eta |> 
    dplyr::filter(Year >= 1971, 
                  ProductAggregation == "Grouped", 
                  IndustryAggregation == "Grouped", 
                  GrossNet == "Gross", 
                  IncludesNEU == FALSE) |> 
    dplyr::select(-Final, -Useful)
) |>
  dplyr::mutate(
    Dataset = factor(Dataset, 
                     levels = c("IEA", "MW", "IEA + MW")), 
    Sector = dplyr::case_when(
      Sector == "Agriculture, forestry, and fishing" ~ "Agriculture, forestry,\nand fishing",
      Sector == "Commercial and public services" ~ "Commercial and\npublic services",
      Sector == "Final consumption not elsewhere specified" ~ "Final consumption not\nelsewhere specified",
      Sector == "Energy industry own use" ~ "Energy industry\nown use",
      TRUE ~ .data[["Sector"]]
    ),
    Sector = factor(Sector,
                    levels = c("Energy industry\nown use",
                               "Non-energy use",
                               "Final consumption not\nelsewhere specified",
                               "Transport",
                               "Commercial and\npublic services",
                               "Industry",
                               "Residential",
                               "Agriculture, forestry,\nand fishing",
                               "World"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = etafu, 
                                         colour = Sector)) + 
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Dataset), 
                      cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(name = NULL, 
                              breaks = c(1980, 2000, 2020)) + 
  ggplot2::scale_y_continuous(name = expression("Final-to-useful efficiency ("*eta[fu]*") [%]"),
                              limits = c(0, NA),
                              breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  ggplot2::scale_colour_manual(name = NULL, 
                               values = MKHthemes::sector_colours) +
  
  
  MKHthemes::xy_theme()
```


Figure \@ref(fig:fig12) is a recreation of Figure 12
of @Brockway:2024ab.

```{r fig12, fig.align='center', fig.retina=2, fig.height=5, echo=FALSE, fig.cap=paste0('Gross aggregate efficiencies by continent including both IEA and muscle work energy data (IEA + MW). Non-energy uses are excluded. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 12 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).')} 
world_continent_eta |> 
  dplyr::filter(Country != "GRC", 
                Year >= 1971, 
                Dataset == "IEA + MW",
                GrossNet == "Gross",
                LastStage == "Useful",
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified",
                IncludesNEU == FALSE) |> 
  dplyr::select(-EXp, -EXf, -EXu) |> 
  tidyr::pivot_longer(cols = c("etapf", "etafu", "etapu"), 
                      names_to = "varname", 
                      values_to = "value") |>
  dplyr::mutate(
    Country = dplyr::case_when(
      Country == "MidEast" ~ "Middle East", 
      Country == "NoAmr" ~ "North America",
      Country == "SoCeAmr" ~ "South and\nCentral America",
      TRUE ~ .data[["Country"]]
    ), 
    varname = dplyr::case_when(
      varname == "etapf" ~ "eta[pf]", 
      varname == "etafu" ~ "eta[fu]", 
      varname == "etapu" ~ "eta[pu]", 
      TRUE ~ NA_character_
    ), 
    varname = factor(varname, 
                     levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = value, 
                                         colour = Country)) + 
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(varname),
                      cols = ggplot2::vars(EnergyType), 
                      labeller = ggplot2::label_parsed) + 
  ggplot2::scale_x_continuous(name = NULL, breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = expression("Energy and exergy efficiency ("*eta*") [%]"),
                              limits = c(0, 1),
                              breaks = c(0.0, 0.2, 0.4, 0.6, 0.8, 1),
                              labels = scales::label_percent(suffix = "")) +
  ggplot2::scale_colour_manual(name = NULL, 
                               values = MKHthemes::continent_colours) + 
  
  MKHthemes::xy_theme()
```

Figure \@ref(fig:fig13) is a recreation of Figure 13
of @Brockway:2024ab.

```{r fig13, fig.align='center', fig.retina=2, fig.height=5, echo=FALSE, fig.cap=paste0('Indexed world primary, final, and useful energy and exergy consumption from Figure @fig8. Non-energy uses are excluded. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 13 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).')} 
fig8_data |> 
  matsindf::group_by_everything_except("value", "Year") |> 
  dplyr::mutate(
    value = value / value[Year == 1971],
    varname = factor(varname, 
                     levels = c("Useful", "Primary", "Final"))
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year,
                                         y = value, 
                                         colour = varname, 
                                         linetype = GrossNet)) + 
  ggplot2::geom_hline(yintercept = 1, colour = "gray", linewidth = 0.2) +
  ggplot2::geom_line() + 
  ggplot2::facet_grid(rows = ggplot2::vars(Dataset),
                      cols = ggplot2::vars(EnergyType)) +
  ggplot2::scale_x_continuous(name = NULL, breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(name = "Indexed energy and exergy consumption (1971 = 1)",
                              limits = c(0.7, 4),
                              breaks = c(1, 2, 3, 4)) +
  viridis::scale_color_viridis(name = NULL, direction = -1, discrete = TRUE, end = 0.7) +
  ggplot2::scale_linetype_manual(name = NULL,
                                 values = c(Gross = "solid", Net = "dashed")) +
  MKHthemes::xy_theme()
```

Figure \@ref(fig:fig16) is a recreation of Figure 16
of @Brockway:2024ab.

```{r CoverageStats, echo=FALSE}
calc_db_coverage_stats <- function(completed_allocation_tables,
                                   completed_efficiency_tables,
                                   country = IEATools::iea_cols$country,
                                   c_source = IEATools::template_cols$c_source,
                                   c_source_category = "CSourceCategory",
                                   eta_fu_source = IEATools::template_cols$eta_fu_source,
                                   eta_source_category = "EtaSourceCategory",
                                   own_country = "Own country",
                                   exemplar = "Exemplar",
                                   continent = "Continent",
                                   continents = c(PFUPipelineTools::all_countries$afri,
                                                  PFUPipelineTools::all_countries$asia,
                                                  PFUPipelineTools::all_countries$eurp,
                                                  PFUPipelineTools::all_countries$mide,
                                                  PFUPipelineTools::all_countries$namr,
                                                  PFUPipelineTools::all_countries$ocen,
                                                  PFUPipelineTools::all_countries$samr),
                                   wrld = PFUPipelineTools::all_countries$wrld,
                                   unknown = "Unknown") {

  allocations_with_categories <- completed_allocation_tables |>
    dplyr::mutate(
      "{c_source_category}" := dplyr::case_when(
        .data[[c_source]] == .data[[country]] ~ own_country,
        .data[[c_source]] == wrld ~ wrld,
        .data[[c_source]] %in% continents ~ continent,
        .data[[c_source]] %in% PFUPipelineTools::all_countries ~ exemplar,
        TRUE ~ unknown
      )
    )

  allocation_counts <- dplyr::bind_rows(
    # counts by country
    dplyr::count(dplyr::group_by(allocations_with_categories, .data[[country]]), .data[[c_source_category]]),
    # counts for the world
    dplyr::count(allocations_with_categories, .data[[c_source_category]]) |>
      dplyr::mutate(
        Country = "World"
      )
    ) |>
    tidyr::pivot_wider(names_from = dplyr::all_of(c_source_category), values_from = n) |>
    tidyr::replace_na(replace = list(WRLD = 0, Exemplar = 0, `Own country` = 0)) |>
    dplyr::mutate(
      table = "CompletedAllocationTables"
    )

  efficiencies_with_categories <- completed_efficiency_tables |>
    dplyr::mutate(
      "{eta_source_category}" := dplyr::case_when(
        .data[[eta_fu_source]] == .data[[country]] ~ own_country,
        .data[[eta_fu_source]] == wrld ~ wrld,
        .data[[eta_fu_source]] %in% continents ~ continent,
        .data[[eta_fu_source]] %in% PFUPipelineTools::all_countries ~ exemplar,
        TRUE ~ unknown
      )
    )

  efficiency_counts <- dplyr::bind_rows(
    # counts by country
    dplyr::count(dplyr::group_by(efficiencies_with_categories, .data[[country]]), .data[[eta_source_category]]),
    # counts for the world
    dplyr::count(efficiencies_with_categories, .data[[eta_source_category]]) |>
      dplyr::mutate(
        Country = "World"
      )
    ) |>
    tidyr::pivot_wider(names_from = dplyr::all_of(eta_source_category), values_from = n) |>
    tidyr::replace_na(replace = list(WRLD = 0, Continent = 0, Exemplar = 0, `Own country` = 0)) |>
    dplyr::mutate(
      table = "CompletedEfficiencyTables"
    )

  dplyr::bind_rows(allocation_counts, efficiency_counts) |>
    tidyr::replace_na(replace = list(WRLD = 0, Continent = 0, Exemplar = 0, `Own country` = 0))
}

coverage <- calc_db_coverage_stats(completed_allocation_tables = completed_allocation_tables,
                                   completed_efficiency_tables = completed_efficiency_tables)
```


```{r fig16, fig.align='center', fig.retina=2, fig.height=3, echo=FALSE, fig.cap=paste0('Final-to-useful (FU) machine allocation counts for selected countries (USA, Canada, Great Britain, South Africa, and Ghana) and the World. "World" is the sum of all countries in the CL-PFU database. "WRLD" is a "country" with all possible FU machine allocations, which provides fallback allocations when own-country and exemplar country allocations are unavailable. Note different vertical scales for countries and the World. This is a re-creation (using CL-PFU database ', clpfu_version, ') of Figure 16 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).')} 
coverage |>
  dplyr::filter(Country %in% c("USA", "GBR", "GHA", "ZAF", "World", "CAN"),
                table == "CompletedAllocationTables") |>
  dplyr::mutate(
    Continent = NULL
  ) |>
  tidyr::pivot_longer(cols = c(WRLD, Exemplar, `Own country`),
                      names_to = "var",
                      values_to = "val") |>
  dplyr::mutate(
    which_facet = dplyr::case_when(
      Country == "World" ~ "World",
      TRUE ~ "Countries"
    ),
    table = dplyr::case_when(
      table == "CompletedAllocationTables" ~ "Allocation",
      table == "CompletedEfficiencyTables" ~ "Efficiency",
      TRUE ~ "Unknown"
    ),
    var = factor(var, levels = c("WRLD", "Exemplar", "Own country")),
    Country = factor(Country, levels = c("USA", "CAN", "GBR",
                                         "ZAF", "GHA", "World"))
  ) |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = Country, y = val, fill = var)) +
  ggplot2::geom_col() +
  ggh4x::facet_grid2(cols = ggplot2::vars(which_facet),
                     independent = "y",
                     scales = "free",
                     space = "free_x") +
  ggplot2::scale_x_discrete(name = NULL) +
  ggplot2::scale_y_continuous(name = "FU machine allocation counts", 
                              # breaks = c(0, 5000, 10000, 15000, 20000, 25000),
                              labels = scales::label_number(suffix = "k", scale = 1e-3, big.mark = ","),
                              expand = ggplot2::expansion(mult = c(0, 0.02))) +
  ggplot2::scale_fill_viridis_d(name = NULL,
                                direction = -1,
                                begin = 0.2,
                                end = 0.8) +
  MKHthemes::xy_theme() +
  ggplot2::theme(axis.ticks.length.x = unit(0, "pt"),
                 strip.background = ggplot2::element_blank(),
                 strip.text.x = ggplot2::element_blank())
```


Figure \@ref(fig:GRC-MW) provides an assessment of muscle work for Greece.
Previously, data from the UN's Food and Agriculture Organization (FAO)
had a large jump. 
The large jump was fixed a few years ago. 
The fact that there a large jump is absent in Figure \@ref(fig:GRC-MW)
indicates that we are correctly using the 
fixed FAO data for `r clpfu_version` of the CL-PFU database.

```{r GRC-MW, fig.align='center', fig.retina=2, fig.height=3, echo=FALSE, fig.cap="Greek muscle work."}
world_continent_eta |> 
  dplyr::filter(Dataset == "MW", 
                Country == "GRC", 
                LastStage == "Useful", 
                IncludesNEU == TRUE, 
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified") |> 
  dplyr::mutate(
    etapf = NULL, 
    etafu = NULL, 
    etapu = NULL
  ) |>
  tidyr::pivot_longer(cols = c(EXp, EXf, EXu), 
                      names_to = "variable", 
                      values_to = "value") |> 
  dplyr::mutate(
    variable = dplyr::case_when(
      variable == "EXp" ~ "Primary", 
      variable == "EXf" ~ "Final", 
      variable == "EXu" ~ "Useful", 
      TRUE ~ NA_character_
    ),
    variable = factor(variable, levels = c("Primary", "Final", "Useful")),
    # Convert TJ to EJ
    value = value / 1e6
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = value, 
                                         linetype = GrossNet)) + 
  ggplot2::geom_line() +
  ggplot2::facet_grid(rows = ggplot2::vars(variable),
                        cols = ggplot2::vars(EnergyType)) +
  ggplot2::scale_x_continuous(name = NULL) +
  ggplot2::scale_y_continuous(name = "MW energy and exergy for Greece [EJ]") + 
  ggplot2::labs(linetype = NULL) +
  MKHthemes::xy_theme()
```



Table \@ref(tab:tab4) is a recreation of Table 4
of @Brockway:2024ab.


```{r tab4, echo=FALSE, tab.cap = paste0('Gross world efficiencies including both IEA and muscle work data. Non-energy uses are excluded. This is a recreation (using CL-PFU database ', clpfu_version, ') of Table 4 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).')}
world_continent_eta |> 
  dplyr::filter(Dataset == "IEA + MW", 
                Country == "World", 
                Year %in% c(1971, 2020), 
                LastStage == "Useful",
                IncludesNEU == FALSE, 
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified", 
                GrossNet == "Gross") |>
  dplyr::select(-EXp, -EXf, -EXu) |>
  tidyr::pivot_longer(cols = c("etapf", "etafu", "etapu"),
                      names_to = "variable",
                      values_to = "value") |>
  dplyr::mutate(
    coltitle = paste0(EnergyType, "-", Year), 
    EnergyType = NULL, 
    Year = NULL, 
    variable = dplyr::case_when(
      variable == "etapf" ~ "eta[pf]", 
      variable == "etafu" ~ "eta[fu]", 
      variable == "etapu" ~ "eta[pu]", 
      TRUE ~ NA_character_
    )
  ) |> 
  tidyr::pivot_wider(names_from = "coltitle", values_from = "value") |>
  dplyr::select("variable", 
                "Energy-1971", "Energy-2020", 
                "Exergy-1971", "Exergy-2020") |>  
  dplyr::mutate(
    variable = factor(variable, 
                      levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  ) |> 
  dplyr::mutate(
    # Convert to %
    dplyr::across(where(is.numeric),
                  ~ sprintf("%.2f %%", .x * 100)
                  )
  ) |>
  knitr::kable(format = "latex", booktabs = TRUE)
```

Table \@ref(tab:tab5) is a recreation of Table 5
of @Brockway:2024ab.

```{r tab5, tab.cap=paste0('Gross world primary, final, and useful energy and exergy growth metrics. Both IEA and muscle work data are included. Non-energy uses are excluded. This is a recreation (using CL-PFU database ', clpfu_version, ') of Table 5 of \\cite{Brockway:2024ab} (which used CL-PFU database v1.2).'), echo=FALSE}
world_continent_eta |> 
  dplyr::filter(Dataset == "IEA + MW", 
                Country == "World", 
                Year %in% c(1971, 2020), 
                LastStage == "Useful",
                IncludesNEU == FALSE, 
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified", 
                GrossNet == "Gross") |>
  dplyr::select(-etapf, -etafu, -etapu) |> 
  tidyr::pivot_longer(cols = c("EXp", "EXf", "EXu"),
                      names_to = "variable",
                      values_to = "value") |>
  dplyr::mutate(
    coltitle = paste0(EnergyType, "-", Year), 
    EnergyType = NULL, 
    Year = NULL, 
  ) |>
  tidyr::pivot_wider(names_from = "coltitle", values_from = "value") |>
  dplyr::select("variable", 
                "Energy-1971", "Energy-2020", 
                "Exergy-1971", "Exergy-2020") |> 
  dplyr::mutate(
    variable = dplyr::case_when(
      variable == "EXp" ~ "Primary", 
      variable == "EXf" ~ "Final", 
      variable == "EXu" ~ "Useful", 
      TRUE ~ NA_character_
    ), 
    `Energy 2020/1971` = sprintf(fmt = "%.2f x", 
                                 `Energy-2020` / `Energy-1971`), 
    `Exergy 2020/1971` = sprintf(fmt = "%.2f x", 
                                 `Exergy-2020` / `Exergy-1971`),
    `r Energy 2020/1971` = sprintf(fmt = "%.2f %%/yr", 
                                    ((`Energy-2020`/`Energy-1971`) ^ 
                                       (1/(2020-1971)) - 1) * 100),
    `r Exergy 2020/1971` = sprintf(fmt = "%.2f %%/yr",
                                    ((`Exergy-2020`/`Exergy-1971`) ^
                                       (1/(2020-1971)) - 1) * 100), 
    `Energy-1971` = NULL, 
    `Energy-2020` = NULL, 
    `Exergy-1971` = NULL, 
    `Exergy-2020` = NULL
  ) |> 
  knitr::kable(format = "latex", booktabs = TRUE)
```



Figure \@ref(fig:NG-Pipeline-Compressors) demonstrates that we now include
a new machine (Natural gas pipeline compressors) suggested by IST.

```{r NG-Pipeline-Compressors, fig.align='center', fig.retina=2, fig.height=3, echo=FALSE, fig.cap="Enegy use by Natural gas pipeline gas compressors."}
world_mats |> 
  dplyr::filter(Dataset == "IEA", 
                LastStage == "Useful",
                EnergyType == "Energy",
                IncludesNEU == FALSE,
                ProductAggregation == "Specified",
                IndustryAggregation == "Specified") |> 
  dplyr::select(-R, -U, -U_EIOU, -r_EIOU, -V, -Y, -S_units) |> 
  dplyr::mutate(
    U_feed = matsbyname::select_cols_byname(U_feed,
                                            retain_pattern = "^Natural gas pipeline gas compressors") |> 
      matsbyname::clean_byname() |> 
      matsbyname::quotient_byname(1e6) # Convert TJ to EJ
  ) |> 
  tidyr::pivot_longer(cols = "U_feed", names_to = "matnames", values_to = "matvals") |> 
  matsindf::expand_to_tidy() |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, y = matvals, fill = rownames)) + 
  ggplot2::geom_area() + 
  
  ggplot2::scale_x_continuous(name = NULL) +
  ggplot2::scale_y_continuous(name = "Energy use [EJ]",
                              breaks = c(0, 5, 10, 15)) +
  ggplot2::scale_fill_viridis_d(name = NULL, 
                                direction = -1) +
  
  MKHthemes::xy_theme() + 
  ggplot2::theme(
    legend.text = element_text(size = 6), 
    legend.key.size = unit(0.4, "cm")
  )
```


# References
