---
title: "MechWork.Rmd"
output: pdf_document
date: "2026-03-17"
author: Matthew Kuperus Heun
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(matsindf)
library(MKHthemes)
```

```{r colors, include=FALSE}
useful_carrier_colours <- c(AnMech = "purple1", 
                            `Animal mechanical work` = "purple1", 
                            `Animal mechanical\nwork` = "purple1",
                            HuMech = "purple4",
                            `Human mechanical work` = "purple4",
                            `Human mechanical\nwork` = "purple4",
                            MechW = "pink3",
                            `Mechanical work` = "pink3",
                            Prop = "pink1",
                            Propulsion = "pink1",
                            HPF = "ivory3",
                            `High-pressure fluids` = "ivory3",
                            `High-pressure\nfluids` = "ivory3",
                            LTH = "orangered1", 
                            `Low-temperature heat` = "orangered1",
                            `Low-temperature\nheat` = "orangered1",
                            MTH = "orangered3",
                            `Medium-temperature heat` = "orangered3",
                            `Medium-temperature\nheat` = "orangered3",
                            HTH = "orangered4",
                            `High-temperature heat` = "orangered4",
                            `High-temperature\nheat` = "orangered4",
                            LTC = "deepskyblue3",
                            `Low-temperature cooling` = "deepskyblue3",
                            `Low-temperature\ncooling` = "deepskyblue3",
                            IP = "palegreen4",
                            `Information processing` = "palegreen4",
                            `Information\nprocessing` = "palegreen4",
                            L = "gold2",
                            Light = "gold2")
```

```{r stuff-to-exclude, include=FALSE} 
sectors_to_exclude <- RCLabels::make_or_pattern(
  c("Exports", 
    "Exports to World aviation bunkers", 
    "Exports to World marine bunkers", 
    "Losses", 
    "Statistical differences", 
    "Stock changes"))
```



## Introduction

This file investigates questions about
world aggregate mechanical work.
The questions were raised by Tiago Domingos
regarding Figure 7 of the paper
P. E. Brockway, M. K. Heun, Z. Marshall, E. Aramendia, P. Steenwyk, T. Relph, M. Widjanarko,
J. J. Kim, A. Sainju, and J. Irtube. A country-level primary-final-useful (CL-PFU) energy and exergy
database: Overview of its construction and 1971–2020 world-level efficiency results. Environmental
Research: Energy, 1(025005):1–29, jun 2024.

## Data

Gather and save data from the database.

```{r data, eval=FALSE}
conn <- PFUPipelineTools::get_mexerdb_conn(user = "dbcreator")
world_mats <- PFUPipelineTools::pl_filter_collect(
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)
world_mats |>
  saveRDS(file = "world_mats.rds")
```

## Analysis

Load data.

```{r load}
world_mats <- readRDS(file = "world_mats.rds")
```

Make a graph similar to Figure 7
using pre-aggregated (Grouped) data.

```{r}
world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::select(-R, -U, -U_EIOU, -U_feed, -r_EIOU, -V, -S_units) |>
  dplyr::mutate(
    Y = Y |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)  |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU") |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |> 
  tidyr::pivot_longer(cols = "Y", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  dplyr::mutate(
    rownames = factor(rownames, levels = c("IP", "L", "HTH", "MTH", "LTH",
                                           "LTC", "HPF", "Prop", "MechW", "HuMech", "AnMech"))

  ) |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_fill_manual(values = useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme()
```

Aggregate using Specified data to confirm similarity 
of Specified data to Figure 7.

```{r self-aggregation}
# Read useful product aggregation map
eu_product_aggregation <- openxlsx2::read_xlsx("aggregation_mapping.xlsx")
foo <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Despecified", 
                IndustryAggregation == "Despecified") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::select(-R, -U, -U_EIOU, -U_feed, -r_EIOU, -V, -S_units) |> 
  dplyr::mutate(
    Y = Y |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)   )

```
