---
title: "Mechanical work analysis for the CL-PFU database"
output:
  bookdown::pdf_document2:
    toc: false
date: "2026-04-02"
author: Matthew Kuperus Heun
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(Matrix)
library(matsindf)
library(MKHthemes)
library(tidyr)
```

```{r colors, include=FALSE}
useful_carrier_colours <- c(AnMech = "purple1", 
                            `Animal mechanical work` = "purple1", 
                            `Animal mechanical\nwork` = "purple1",
                            HuMech = "purple4",
                            `Human mechanical work` = "purple4",
                            `Human mechanical\nwork` = "purple4",
                            MechW = "pink3",
                            `Mechanical work` = "pink3",
                            Prop = "pink1",
                            Propulsion = "pink1",
                            HPF = "ivory3",
                            `High-pressure fluids` = "ivory3",
                            `High-pressure\nfluids` = "ivory3",
                            LTH = "orangered1", 
                            `Low-temperature heat` = "orangered1",
                            `Low-temperature\nheat` = "orangered1",
                            MTH = "orangered3",
                            `Medium-temperature heat` = "orangered3",
                            `Medium-temperature\nheat` = "orangered3",
                            HTH = "orangered4",
                            `High-temperature heat` = "orangered4",
                            `High-temperature\nheat` = "orangered4",
                            LTC = "deepskyblue3",
                            `Low-temperature cooling` = "deepskyblue3",
                            `Low-temperature\ncooling` = "deepskyblue3",
                            IP = "palegreen4",
                            `Information processing` = "palegreen4",
                            `Information\nprocessing` = "palegreen4",
                            L = "gold2",
                            Light = "gold2")
```

```{r stuff-to-exclude, include=FALSE} 
sectors_to_exclude <- RCLabels::make_or_pattern(
  c("Exports", 
    "Exports to World aviation bunkers", 
    "Exports to World marine bunkers", 
    "Losses", 
    "Statistical differences", 
    "Stock changes"))
```



# Introduction

This report investigates 
world aggregate Mechanical work.
The questions were raised by Tiago Domingos
regarding Figure 7 of the paper
Brockway et al. A country-level primary-final-useful (CL-PFU) energy and exergy
database: Overview of its construction and 1971–2020 world-level efficiency results. 
Environmental Research: Energy, 1(025005):1–29, jun 2024.
(The "database paper.")


# Data

The first step is to gather relevant data from the database and 
save locally for convenience.

```{r data, eval=FALSE}
conn <- PFUPipelineTools::get_mexerdb_conn(user = "dbcreator")
on.exit(DBI::dbDisconnect(conn))

world_mats <- PFUPipelineTools::pl_filter_collect(
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)
world_mats |>
  saveRDS(file = "world_mats.rds")

DBI::dbDisconnect(conn)
```

```{r load, echo=FALSE}
world_mats <- readRDS(file = "world_mats.rds")
```


# Analysis

## Recreate Figure 7 from the database paper

Note that all analyses are performed at the world level,
with gross useful energy and exergy.
(Gross includes useful stage energy and exergy consumed by final demand
_and_ energy industry own use.
These data are taken from the **Y** and **U**~EIOU~ matrices.)
This analysis employs v2.0 of the CL-PFU database, while 
the paper (and Figure 7) used v1.2.
Some small differences can be expected.

The first step in the analysis is to create Figure \@ref(fig:fig7), 
a graph similar to Figure 7 from the database paper. 
By this step, we double-check the data we obtained from the database.
Here, we're using aggregated (Grouped) data.
We see that Mechanical work (MechW)
appears nearly constant, as observed (and questioned) by Tiago.

```{r fig7, fig.width=6, fig.height=3, fig.align='center', fig.retina=2, fig.cap='Remake of Figure 7 of the database paper using pre-aggregated (i.e., Grouped) data.', echo=FALSE}
grouped_data <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::mutate(
    # For gross useful energy, we need to include both Y and U_EIOU matrices
    YUEIOU = matsbyname::sum_byname(Y, U_EIOU) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)  |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU") |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |>
  dplyr::select(-R, -U, -U_feed, -U_EIOU, -r_EIOU, -V, -Y, -S_units) |>
  tidyr::pivot_longer(cols = "YUEIOU", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  dplyr::mutate(
    rownames = factor(rownames, levels = c("IP", "L", "HTH", "MTH", "LTH",
                                           "LTC", "HPF", "Prop", "MechW", "HuMech", "AnMech"))

  )
grouped_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 250, by = 50)) +
  ggplot2::scale_fill_manual(values = useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme()
```

Next, we create a graph using Specified data (the most raw data)
to confirm equivalence of Specified data and
the Grouped data of Figure \@ref(fig:fig7) (and Figure 7 of the database paper).
The Specified data are required to disaggregate Mechanical work.
Figure \@ref(fig:self-aggregation-graph) is the same as 
Figure \@ref(fig:fig7), indicating 
(a) that we have consistency in our data 
and 
(b) that we can use the Specified data to disaggregate
    the Mechanical work time series.

```{r self-aggregation-graph, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Remake of Figure 7 from the database paper starting from Specified data.', echo=FALSE}
# Read useful product aggregation map
eu_product_agg_map <- openxlsx2::read_xlsx("aggregation_mapping.xlsx", sheet = "eu_product_aggregation") |> 
  matsbyname::agg_table_to_agg_map(few_colname = "Few", many_colname = "Many")
nouns <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::select(-R, -U, -U_feed, -r_EIOU, -V, -S_units)

aggregated_nouns <- nouns |> 
  dplyr::mutate(
    YUEIOU = matsbyname::sum_byname(Y, U_EIOU) |> 
      matsbyname::aggregate_pieces_byname(piece = "noun", 
                                          margin = 1,
                                          notation = RCLabels::bracket_notation) |> 
      matsbyname::aggregate_pieces_byname(piece = "noun", 
                                          margin = 2, 
                                          notation = RCLabels::bracket_notation) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude) |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU")
  ) |> 
  dplyr::select(-U_EIOU, -Y)
      
grouped <- aggregated_nouns |>       
  dplyr::mutate(
    YUEIOU = YUEIOU |> 
      matsbyname::aggregate_byname(aggregation_map = list(eu_product_agg_map), 
                                   margin = 1) |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |> 
  tidyr::pivot_longer(cols = "YUEIOU", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |> 
  dplyr::mutate(
    rownames = factor(rownames, levels = c("IP", "L", "HTH", "MTH", "LTH",
                                           "LTC", "HPF", "Prop", "MechW", "HuMech", "AnMech"))
    
  )

grouped |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 250, by = 50)) +
  ggplot2::scale_fill_manual(values = useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme()
```


Next, we focus on Mechanical work.


## Mechanical work

```{r, echo=FALSE}
# Mechanical work numbers
grouped_data_1971_2020 <- grouped_data |> 
  dplyr::filter(rownames == "MechW", Year %in% c(1971, 2020)) |> 
  dplyr::select(Year, matvals, EnergyType) |> 
  tidyr::pivot_wider(names_from = "EnergyType", values_from = "matvals") 
mech_work_1971 <- grouped_data_1971_2020$E[[1]]
mech_work_2020 <- grouped_data_1971_2020$E[[2]]
perc_inc <- (mech_work_2020 - mech_work_1971) / mech_work_1971 * 100
growth_rate <- ((mech_work_2020/mech_work_1971)^(1/(2020-1971)) - 1) * 100
```

The Mechanical work values for 1971 and 2020 are shown in 
Table \@ref(tab:MechW-table).
Mechanical work is not _actually_ constant over time.
Over the 50-year period, 
Mechanical work grows `r sprintf("%.1f", perc_inc)`% 
for a compound average annual growth rate of
`r sprintf("%.1f", growth_rate)` %/yr.

```{r MechW-table, results='asis', echo=FALSE}
grouped_data_1971_2020 |> 
  xtable::xtable(label = "tab:MechW-table", 
                 caption = "Mechanical work values (in EJ) for 1971 and 2000.") |> 
  print(comment = FALSE, 
        caption.placement = "top", 
        include.rownames = FALSE, 
        booktabs = TRUE)
```

The Mechanical work time series are shown in Figure \@ref(fig:mech-work-only).

```{r mech-work-only, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='World mechanical work.', echo=FALSE}
grouped_data |> 
  dplyr::filter(rownames == "MechW") |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, y = matvals)) + 
  ggplot2::geom_line() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(limits = c(0, NA)) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Mechanical work (MechW) [EJ]") +
  MKHthemes::xy_theme()
```

Next, we disaggregate Mechanical work into subcategories.
In the CL-PFU database, we calculate aggregate Mechanical work by summing
`r paste(paste(eu_product_agg_map$MechW[1:3], collapse = ", "), eu_product_agg_map$MechW[4], sep = ", and ")`.
Figure \@ref(fig:mech-work) shows that Mechanical work (MechW) 
is growing (in total), 
but Mechanical drive (MD) is declining slightly
while Machine work (MacW) and Kinetic energy (KE)
grow through time.
Material fracture (MF) is insignificant 
over the timespan of the CL-PFU database.

```{r mech-work, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Disaggregation of mechanical work.', echo=FALSE}
mech_work_pattern <- RCLabels::make_or_pattern(eu_product_agg_map$MechW)
mechw_details_data <- aggregated_nouns |> 
  dplyr::mutate(
    YUEIOU = YUEIOU |> 
      matsbyname::select_rows_byname(retain_pattern = mech_work_pattern) |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  )  |> 
  tidyr::pivot_longer(cols = "YUEIOU", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) 

mechw_graph <- mechw_details_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_fill_manual(values = c(KE = "rosybrown2", 
                                        MacW = "pink3", 
                                        MD = "rosybrown4", 
                                        MF = "black")) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Mecahnical work components [EJ]") +
  MKHthemes::xy_theme()
mechw_graph
```

Now, we plot disaggregated Mechanical work at the same scale as Figure 7
from the database paper.
Here, we see that Mechanical work appears quite flat, 
as Tiago noticed, despite its growth.

```{r mech-work-full-scale, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Mechanical work at same scale as Figure 7 from the database paper.', echo=FALSE}
mechw_graph +
  ggplot2::scale_y_continuous(limits = c(0, 280), 
                              breaks = seq(0, 250, by = 50))
```

Table \@ref(tab:mechw-disagg-table) shows 
numerical values for disaggregated Mechanical work in 1971 and 2020.

```{r mechw-disagg-table, results='asis', tab.cap="Mechanical work components (in EJ) in 1971 and 2000.", echo=FALSE}
mechw_details_data |> 
  dplyr::filter(Year %in% c(1971, 2020), EnergyType == "E") |> 
  dplyr::select(Year, rownames, matvals) |> 
  tidyr::pivot_wider(names_from = "rownames", values_from = "matvals") |> 
  xtable::xtable(label = "tab:mechw-disagg-table", 
                 caption = "Mechanical work components (in EJ) for 1971 and 2000.") |> 
  print(comment = FALSE, 
        caption.placement = "top", 
        include.rownames = FALSE, 
        booktabs = TRUE)
```


## Mechanical drive (MD)

```{r md-calcs, echo=FALSE}
md_mats <- world_mats |> 
  dplyr::filter(Year >= 1971, 
                Dataset == "CL-PFU IEA+MW",
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified") |> 
  dplyr::mutate(
    U_MD = U |> 
      matsbyname::select_rowcol_piece_byname(retain = "MD", 
                                             piece = "suff", 
                                             notation = RCLabels::arrow_notation, 
                                             margin = 2) |> 
      matsbyname::clean_byname(margin = 1) |> 
      matsbyname::colsums_byname(rowname = "Input") |> 
      matsbyname::transpose_byname(),
    V_MD = V |> 
      matsbyname::select_rowcol_piece_byname(retain = "MD", 
                                             piece = "suff",
                                             notation = RCLabels::arrow_notation, 
                                             margin = 1) |> 
      matsbyname::clean_byname(margin = 2) |> 
      matsbyname::rowsums_byname(colname = "Output")
  ) |> 
  dplyr::select(-R, -U, -U_EIOU, -U_feed, -r_EIOU, -V, -Y, -S_units)


md_df <- md_mats |> 
  tidyr::pivot_longer(cols = c("U_MD", "V_MD"),
                      names_to = "matnames", 
                      values_to = "matvals") |> 
  matsindf::expand_to_tidy(drop = 0) |> 
  dplyr::mutate(
    machine = rownames |> 
      RCLabels::get_nouns(notation = RCLabels::arrow_notation), 
    direction = colnames,
    matnames = NULL, 
    rownames = NULL, 
    colnames = NULL, 
    rowtypes = NULL, 
    coltypes = NULL
  ) |> 
  tidyr::pivot_wider(names_from = "direction", values_from = "matvals")

md_machines <- unique(md_df$machine)
md_machines_str <- paste(paste(md_machines[1:2],
                               collapse = ", "),
                         md_machines[3], 
                         sep = ", and ")
```

The largest subcategory of Mechanical work (MechW) is 
Mechanical drive (MD), 
and MD is quite flat over time. 
To gain further insight, 
we extract, 
from the **V** (make) matrix when the last stage is Useful,
machines that produce Mechanical drive (MD).
The machines that produce MD are
`r md_machines_str`.
Figure \@ref(fig:MD-components) shows 
the evolution of world-total outputs 
from those three machines. 

```{r MD-components, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='World mechanical drive (MD) by source.', echo=FALSE}
md_df |> 
  dplyr::mutate(
    Output = Output / 1e6 # Convert TJ to EJ
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, y = Output, fill = machine)) +
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 10, by = 2)) +
  ggplot2::scale_fill_manual(
    values = c(`Domestic appliances` = "steelblue1", 
               `Static diesel engines` = "steelblue3", 
               `Static gasoline engines` = "steelblue4")) +
  ggplot2::labs(x = NULL, y = "World MD [EJ]", fill = NULL) +
  MKHthemes::xy_theme()
```

Figure \@ref(fig:MD-components) shows 
that Static diesel engines are producing less MD over time, 
a decline that is partially offset by growing
MD production by Domestic appliances.
Static gasoline engines provide a very minor contribution 
to total Mechanical drive.


## Inputs to MD machines

`r md_machines_str` consume energy to make MD. 
We extract, 
from the **U** (use) matrix when the last stage is Useful,
energy and exergy inputs 
to machines that make Mechanical drive (MD).
Figure \@ref(fig:MD-inputs) shows 
the evolution of world-total inputs
to those machines. 

```{r MD-inputs, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='World inputs for mechanical drive (MD) by machine.', echo=FALSE}
md_df |> 
  dplyr::mutate(
    Input = Input / 1e6 # Convert TJ to EJ
  ) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, y = Input, fill = machine)) +
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 30, by = 5)) +
  ggplot2::scale_fill_manual(
    values = c(`Domestic appliances` = "steelblue1", 
               `Static diesel engines` = "steelblue3", 
               `Static gasoline engines` = "steelblue4")) +
  ggplot2::labs(x = NULL, y = "World inputs for MD [EJ]", fill = NULL) +
  MKHthemes::xy_theme()
```

Figure \@ref(fig:MD-inputs) shows that input energy
for Static diesel engines is declining over time. 
With declining MD output from Static diesel engines, 
these results suggest declining usage of 
Static diesel engines over time.

Both input energy to and output energy from
Domestic appliances is growing over time,
suggesting growing use of Domestic appliances over time.


## Efficiencies of MD machines

It is also instructive to look at the efficiency of the machines that 
produce MD
(`r md_machines_str`).
Efficiency can be calculated by dividing MD machine outputs
(found in the **V** matrix) by MD machine inputs
(found in the **U** matrix).
Figure \@ref(fig:MD-efficiencies) shows 
the evolution of those 
efficiencies.

```{r MD-efficiencies, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='World-level efficiencies of machines that produce MD.', echo=FALSE}
md_df |> 
  dplyr::mutate(eta = Output / Input) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, y = eta, linetype = machine, colour = machine)) +
  ggplot2::geom_line() +
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  ggplot2::scale_colour_manual(
    values = c(`Domestic appliances` = "steelblue1", 
               `Static diesel engines` = "steelblue3", 
               `Static gasoline engines` = "steelblue4")) +
  ggplot2::scale_linetype_manual(
    values = c(`Domestic appliances` = 1, 
               `Static diesel engines` = 2, 
               `Static gasoline engines` = 3)) +
  ggplot2::labs(x = NULL, 
                y = "WRLD MD machine efficiencies [-]", 
                linetype = NULL, 
                colour = NULL) +
  MKHthemes::xy_theme()
```


A few observations can be made.

* There is an interesting inflection in World engine efficiencies
  around 2000 that should be investigated.
* The efficiency of Static diesel engines rises 
  from 1971--2020, thus
  MD output declines more slowly than energy inputs. 
* The energy and exergy 
  output from (i.e., produced by)
  MD machines is identical.
  MD machine efficiencies are identical across energy and exergy for 
  Domestic appliances, because 
  the energy and exergy of input electricity is identical.
  Most Domestic appliances are powered by Electricity.
* The efficiency of Static diesel and gasoline engines
  is slightly different between energy and exergy
  due to the slight difference in
  energy and exergy of input fossil fuels.
  Because the exergy is larger than exergy for fossil fuels, 
  exergy efficiency is slightly lower for 
  Static diesel and gasoline engines.


## Consumption of MD by sectors

We can interrogate the **Y** (final demand) 
matrix for usage of MD by various sectors.
Figure \@ref(fig:md-Y-graph) shows 
the evolution of MD consumption.

```{r MD-Y, echo=FALSE}
md_Y <- world_mats |> 
  dplyr::filter(Year >= 1971, 
                Dataset == "CL-PFU IEA+MW",
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified") |> 
  dplyr::select(-R, -U, -U_feed, -U_EIOU, -r_EIOU, -V, -S_units) |>
  dplyr::mutate(
    Y_MD = Y |> 
      matsbyname::select_rowcol_piece_byname(retain = "MD", 
                                             piece = "pref", 
                                             notation = RCLabels::from_notation, 
                                             margin = 1) |> 
      matsbyname::clean_byname(margin = 2) |> 
      matsbyname::colsums_byname(rowname = "MD input") |> 
      matsbyname::transpose_byname(), 
    Y = NULL
  )


md_Y_df <- md_Y |> 
  tidyr::pivot_longer(cols = Y_MD, 
                      names_to = "matnames", 
                      values_to = "matvals") |> 
  matsindf::expand_to_tidy(drop = 0) |>
  dplyr::rename(
    Sector = rownames
  ) |> 
  dplyr::mutate(
    colnames = NULL, 
    rowtypes = NULL, 
    coltypes = NULL, 
    matnames = NULL, 
    matvals = matvals / 1e6 # Convert TJ to EJ
  )
```


```{r md-Y-graph, fig.retina=2, fig.align='center', fig.cap='World-level MD consumption by sector.', echo=FALSE}
md_Y_df |> 
  dplyr::filter(EnergyType == "X") |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = Sector)) +
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::labs(x = NULL, 
                y = "MD consumption by sector [EJ]", 
                fill = NULL) +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 10, by = 2)) +
  ggplot2::scale_fill_viridis_d(direction = -1, option = "inferno") +
  MKHthemes::xy_theme()
```

The Iron and steel sector is largely decreasing in MD consumption,
while Residential exhibits large increases in MD consumption.

Table \@ref(tab:MDY-table) shows differences
in MD consumption over the 60-year period from 1971 to 2020.

```{r MD-table, results='asis', echo=FALSE}
md_Y_df |> 
  dplyr::filter(Year %in% c(1971, 2020), EnergyType == "X") |> 
  dplyr::select(Year, Sector, matvals) |> 
  tidyr::pivot_wider(names_from = Year, values_from = matvals) |> 
  dplyr::mutate(
    Difference = `2020` - `1971`
  ) |> 
  dplyr::arrange(dplyr::desc(Difference)) |> 
  xtable::xtable(label = "tab:MDY-table", 
                 caption = "Mechanical drive (MD) exergy values (in EJ) for 1971 and 2000, arranged by difference.") |> 
  print(comment = FALSE, 
        caption.placement = "top", 
        include.rownames = FALSE, 
        booktabs = TRUE)

```


# Conclusion

It is true that Mechanical work (MechW) 
grows more slowly than other categories of 
useful energy and exergy.
However, Mechanical work _is growing_
at `r sprintf("%.1f", growth_rate)` %/yr,
about 1/3 of the compound average annual world GDP growth rate
over the same period.

The largest category of MechW is Mechanical drive (MD)
produced by `r md_machines_str`.
The largest producer of MD is Static diesel engines, but
Static diesel engines produce less energy in 2020 than in 1971,
despite rising efficiency, 
causing decreasing energy inputs due to 
less usage over time.
Industrial usage of MD are declining over time,
offsetting increases in consumption of MD by the Residential sector.
