---
title: "MechWork.Rmd"
output: pdf_document
date: "2026-03-17"
author: Matthew Kuperus Heun
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(matsindf)
```


## Introduction

This file investigates questions about
world aggregate mechanical work.
The questions were raised by Tiago Domingos
regarding Figure 7 of the paper
P. E. Brockway, M. K. Heun, Z. Marshall, E. Aramendia, P. Steenwyk, T. Relph, M. Widjanarko,
J. J. Kim, A. Sainju, and J. Irtube. A country-level primary-final-useful (CL-PFU) energy and exergy
database: Overview of its construction and 1971–2020 world-level efficiency results. Environmental
Research: Energy, 1(025005):1–29, jun 2024.

## Data

Gather and save data from the database.

```{r data, eval=FALSE}
conn <- PFUPipelineTools::get_mexerdb_conn(user = "dbcreator")
world_mats <- PFUPipelineTools::pl_filter_collect(
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)
world_mats |>
  saveRDS(file = "world_mats.rds")
```

## Analysis

Load data.

```{r load}
getwd()
world_mats <- readRDS(file = "world_mats.rds")
```


```{r} 
world_mats_filtered <- world_mats |>
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                LastStage == "Useful",
                IncludesNEU == FALSE) |>
  dplyr::select(-R, -U, -U_EIOU, -U_feed, -r_EIOU, -V, -S_units) |>
  dplyr::mutate(
    ProductAggregation = factor(ProductAggregation, levels = c("Specified", "Despecified", "Grouped")), 
    IndustryAggregation = factor(IndustryAggregation, levels  = c("Specified", "Despecified", "Grouped"))
  ) |> 
  dplyr::arrange(Year, ProductAggregation, IndustryAggregation)
```

Make a graph similar to Figure 7.

```{r}
world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::select(-R, -U, -U_EIOU, -U_feed, -r_EIOU, -V, -S_units) |>
  dplyr::mutate(
    ProductAggregation = factor(ProductAggregation, levels = c("Specified", "Despecified", "Grouped")), 
    IndustryAggregation = factor(IndustryAggregation, levels  = c("Specified", "Despecified", "Grouped")),
    Y = Y |> 
      matsbyname::select_cols_byname(
        remove_pattern = RCLabels::make_or_pattern(c("Exports", 
                                                     "Exports to World aviation bunkers", 
                                                     "Exports to World marine bunkers", 
                                                     "Losses", 
                                                     "Statistical differences", 
                                                     "Stock changes"))) |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6)
  ) |> 
  tidyr::pivot_longer(cols = "Y", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType))
```


