---
title: "Mechanical work analysis for the CL-PFU database"
output:
  bookdown::pdf_document2:
    toc: false
date: "2026-03-17"
author: Matthew Kuperus Heun
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(Matrix)
library(matsindf)
library(MKHthemes)
library(tidyr)
```

```{r colors, include=FALSE}
useful_carrier_colours <- c(AnMech = "purple1", 
                            `Animal mechanical work` = "purple1", 
                            `Animal mechanical\nwork` = "purple1",
                            HuMech = "purple4",
                            `Human mechanical work` = "purple4",
                            `Human mechanical\nwork` = "purple4",
                            MechW = "pink3",
                            `Mechanical work` = "pink3",
                            Prop = "pink1",
                            Propulsion = "pink1",
                            HPF = "ivory3",
                            `High-pressure fluids` = "ivory3",
                            `High-pressure\nfluids` = "ivory3",
                            LTH = "orangered1", 
                            `Low-temperature heat` = "orangered1",
                            `Low-temperature\nheat` = "orangered1",
                            MTH = "orangered3",
                            `Medium-temperature heat` = "orangered3",
                            `Medium-temperature\nheat` = "orangered3",
                            HTH = "orangered4",
                            `High-temperature heat` = "orangered4",
                            `High-temperature\nheat` = "orangered4",
                            LTC = "deepskyblue3",
                            `Low-temperature cooling` = "deepskyblue3",
                            `Low-temperature\ncooling` = "deepskyblue3",
                            IP = "palegreen4",
                            `Information processing` = "palegreen4",
                            `Information\nprocessing` = "palegreen4",
                            L = "gold2",
                            Light = "gold2")
```

```{r stuff-to-exclude, include=FALSE} 
sectors_to_exclude <- RCLabels::make_or_pattern(
  c("Exports", 
    "Exports to World aviation bunkers", 
    "Exports to World marine bunkers", 
    "Losses", 
    "Statistical differences", 
    "Stock changes"))
```



# Introduction

This report investigates 
world aggregate Mechanical work.
The questions were raised by Tiago Domingos
regarding Figure 7 of the paper
Brockway et al. A country-level primary-final-useful (CL-PFU) energy and exergy
database: Overview of its construction and 1971–2020 world-level efficiency results. 
Environmental Research: Energy, 1(025005):1–29, jun 2024.


# Data

The first step is to gather relevant data from the database and 
save locally for convenience.

```{r data, eval=FALSE}
conn <- PFUPipelineTools::get_mexerdb_conn(user = "dbcreator")
world_mats <- PFUPipelineTools::pl_filter_collect(
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)
world_mats |>
  saveRDS(file = "world_mats.rds")
```

```{r load, echo=FALSE}
world_mats <- readRDS(file = "world_mats.rds")
```


# Analysis

Note that all analyses are performed at the world level,
with gross useful energy and exergy.
(Gross includes useful stage energy and exergy consumed by final demand
_and_ energy industry own use.
These data are taken from the **Y** and **U_EIOU** matrices.)
This analysis employs v2.0 of the CL-PFU database, while 
the paper (and Figure 7) used v1.2.
Some small differences can be expected.

The first step in the analysis is to create Figure \@ref(fig:fig7), 
a graph similar to Figure 7 from the paper. 
By this step, we double-check the data we obtained from the database.
Here, we're using aggregated (Grouped) data.
We see that Mechanical work (MechW)
appears nearly constant, as observed (and questioned) by Tiago.

```{r fig7, fig.width=6, fig.height=3, fig.align='center', fig.retina=2, fig.cap='Remake of Figure 7 using pre-aggregated (i.e., grouped) data.', echo=FALSE}
grouped_data <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::mutate(
    # For gross useful energy, we need to include both Y and U_EIOU matrices
    YUEIOU = matsbyname::sum_byname(Y, U_EIOU) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)  |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU") |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |>
  dplyr::select(-R, -U, -U_feed, -U_EIOU, -r_EIOU, -V, -Y, -S_units) |>
  tidyr::pivot_longer(cols = "YUEIOU", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  dplyr::mutate(
    rownames = factor(rownames, levels = c("IP", "L", "HTH", "MTH", "LTH",
                                           "LTC", "HPF", "Prop", "MechW", "HuMech", "AnMech"))

  )
grouped_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 250, by = 50)) +
  ggplot2::scale_fill_manual(values = useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme()
```

```{r, echo=FALSE}
# Mechanical work numbers
grouped_data_1971_2020 <- grouped_data |> 
  dplyr::filter(rownames == "MechW", Year %in% c(1971, 2020)) |> 
  dplyr::select(Year, matvals, EnergyType) |> 
  tidyr::pivot_wider(names_from = "EnergyType", values_from = "matvals") 
mech_work_1971 <- grouped_data_1971_2020$E[[1]]
mech_work_2020 <- grouped_data_1971_2020$E[[2]]
perc_inc <- (mech_work_2020 - mech_work_1971) / mech_work_1971 * 100
growth_rate <- ((mech_work_2020/mech_work_1971)^(1/(2020-1971)) - 1) * 100
```

The Mechanical work values for 1971 and 2020 are shown in 
Table \@ref(tab:MechW-table).
Mechanical work is not _actually_ constant over time.
Over the 50-year period, 
Mechanical work grows `r sprintf("%.1f", perc_inc)`% 
for a compound average annual growth rate of
`r sprintf("%.1f", growth_rate)` %/yr.

```{r MechW-table, results='asis', echo=FALSE}
grouped_data_1971_2020 |> 
  xtable::xtable(label = "tab:MechW-table", 
                 caption = "Mechanical work values (in EJ) for 1971 and 2000.") |> 
  print(comment = FALSE, 
        caption.placement = "top", 
        include.rownames = FALSE, 
        booktabs = TRUE)
```

The Mechanical work time series are shown in Figure \@ref(fig:mech-work-only).

```{r mech-work-only, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Mechanical work.', echo=FALSE}
grouped_data |> 
  dplyr::filter(rownames == "MechW") |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, y = matvals)) + 
  ggplot2::geom_line() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(limits = c(0, NA)) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Mechanical work [EJ]") +
  MKHthemes::xy_theme()
```

Next, we create a graph using Specified data (the most raw data)
to confirm equivalance of Specified data and
the Grouped data of Figure \@ref(fig:fig7) (and Figure 7 of the paper).
The Specified data are required to disaggregate Mechanical work.
Figure \@ref(fig:self-aggregation-graph) is the same as 
Figure \@ref(fig:fig7), indicating 
(a) that we have consistency in our data 
and 
(b) that we can use the Specified data to disaggregate
    the Mechanical work time series.

```{r self-aggregation-graph, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Remake of Figure 7 starting from Specified data.', echo=FALSE}
# Read useful product aggregation map
eu_product_agg_map <- openxlsx2::read_xlsx("aggregation_mapping.xlsx", sheet = "eu_product_aggregation") |> 
  matsbyname::agg_table_to_agg_map(few_colname = "Few", many_colname = "Many")
nouns <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::select(-R, -U, -U_feed, -r_EIOU, -V, -S_units)

aggregated_nouns <- nouns |> 
  dplyr::mutate(
    YUEIOU = matsbyname::sum_byname(Y, U_EIOU) |> 
      matsbyname::aggregate_pieces_byname(piece = "noun", 
                                          margin = 1,
                                          notation = RCLabels::bracket_notation) |> 
      matsbyname::aggregate_pieces_byname(piece = "noun", 
                                          margin = 2, 
                                          notation = RCLabels::bracket_notation) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude) |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU")
  ) |> 
  dplyr::select(-U_EIOU, -Y)
      
grouped <- aggregated_nouns |>       
  dplyr::mutate(
    YUEIOU = YUEIOU |> 
      matsbyname::aggregate_byname(aggregation_map = list(eu_product_agg_map), 
                                   margin = 1) |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |> 
  tidyr::pivot_longer(cols = "YUEIOU", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |> 
  dplyr::mutate(
    rownames = factor(rownames, levels = c("IP", "L", "HTH", "MTH", "LTH",
                                           "LTC", "HPF", "Prop", "MechW", "HuMech", "AnMech"))
    
  )

grouped |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 250, by = 50)) +
  ggplot2::scale_fill_manual(values = useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme()
```

Next, we focus on Mechanical work, 
specifically the disaggregation into each category of Mechanical work.
In the CL-PFU database, we calculate aggregate Mechanical work by summing
`r paste(paste(eu_product_agg_map$MechW[1:3], collapse = ", "), eu_product_agg_map$MechW[4], sep = " and ")`.
Figure \@ref(fig:mech-work) shows that Mechanical work (MechW) 
is growing (in total), 
but Mechanical drive (MD) is declining slightly
while Machine work (MacW) and Kinetic energy (KE)
grow through time.
Material fracture (MF) is insignificant 
over the timespan of the CL-PFU database.

```{r mech-work, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Disaggregation of mechanical work.', echo=FALSE}
mech_work_pattern <- RCLabels::make_or_pattern(eu_product_agg_map$MechW)
mechw_details_data <- aggregated_nouns |> 
  dplyr::mutate(
    YUEIOU = YUEIOU |> 
      matsbyname::select_rows_byname(retain_pattern = mech_work_pattern) |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  )  |> 
  tidyr::pivot_longer(cols = "YUEIOU", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) 

mechw_graph <- mechw_details_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_fill_manual(values = c("rosybrown2", "pink3", "rosybrown4", "black")) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Mecahnical work components [EJ]") +
  MKHthemes::xy_theme()
mechw_graph
```

Now, we plot disaggregated Mechanical work at the same scale as Figure 7.
Here, we see that Mechanical work appears quite flat, 
as Tiago noticed, despite its growth.

```{r mech-work-full-scale, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Mechanical work at same scale as Figure 7.', echo=FALSE}
mechw_graph +
  ggplot2::scale_y_continuous(limits = c(0, 280), 
                              breaks = seq(0, 250, by = 50))
```

Table \@ref(tab:mechw-disagg-table) shows 
numerical values for disaggregated Mechanical work in 1971 and 2020.

```{r mechw-disagg-table, results='asis', tab.cap="Mechanical work components (in EJ) in 1971 and 2000.", echo=FALSE}
mechw_details_data |> 
  dplyr::filter(Year %in% c(1971, 2020), EnergyType == "E") |> 
  dplyr::select(Year, rownames, matvals) |> 
  tidyr::pivot_wider(names_from = "rownames", values_from = "matvals") |> 
  xtable::xtable(label = "tab:mechw-disagg-table", 
                 caption = "Mechanical work components (in EJ) for 1971 and 2000.") |> 
  print(comment = FALSE, 
        caption.placement = "top", 
        include.rownames = FALSE, 
        booktabs = TRUE)
```

# Conclusion

It is true that Mechanical work grows more slowly than other categories of 
useful energy and exergy.
However, Mechanical work _is_ growing at `r sprintf("%.1f", growth_rate)` %/yr,
about 1/3 of the compound average annual world GDP growth rate
over the same period.
