---
title: "MechWork.Rmd"
output: pdf_document
date: "2026-03-17"
author: Matthew Kuperus Heun
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This file investigates questions about
world aggregate mechanical work.
The questions were raised by Tiago Domingos
regarding Figure 7 of the paper
P. E. Brockway, M. K. Heun, Z. Marshall, E. Aramendia, P. Steenwyk, T. Relph, M. Widjanarko,
J. J. Kim, A. Sainju, and J. Irtube. A country-level primary-final-useful (CL-PFU) energy and exergy
database: Overview of its construction and 1971–2020 world-level efficiency results. Environmental
Research: Energy, 1(025005):1–29, jun 2024.

## Data

Gather and save data from the database.

```{r data, eval=FALSE}
conn <- PFUPipelineTools::get_mexerdb_conn(user = "dbcreator")
world_mats <- PFUPipelineTools::pl_filter_collect(
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)
world_mats |>
  saveRDS(file = "world_mats.rds")
```

## Analysis

Load data.

```{r load}
getwd()
world_mats <- readRDS(file = "world_mats.rds")
```


```{r} 
world_mats |>
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                EnergyType == "E",
                LastStage == "Useful",
                IncludesNEU == FALSE) |>
  dplyr::select(-R, -U, -U_EIOU, -U_feed, -r_EIOU, -V, -S_units) |>
  dplyr::arrange(Year, ProductAggregation)
```


