---
title: "Mechanical work study"
output:
  bookdown::pdf_document2
date: "2026-03-17"
author: Matthew Kuperus Heun
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(Matrix)
library(matsindf)
library(MKHthemes)
library(tidyr)
```

```{r colors, include=FALSE}
useful_carrier_colours <- c(AnMech = "purple1", 
                            `Animal mechanical work` = "purple1", 
                            `Animal mechanical\nwork` = "purple1",
                            HuMech = "purple4",
                            `Human mechanical work` = "purple4",
                            `Human mechanical\nwork` = "purple4",
                            MechW = "pink3",
                            `Mechanical work` = "pink3",
                            Prop = "pink1",
                            Propulsion = "pink1",
                            HPF = "ivory3",
                            `High-pressure fluids` = "ivory3",
                            `High-pressure\nfluids` = "ivory3",
                            LTH = "orangered1", 
                            `Low-temperature heat` = "orangered1",
                            `Low-temperature\nheat` = "orangered1",
                            MTH = "orangered3",
                            `Medium-temperature heat` = "orangered3",
                            `Medium-temperature\nheat` = "orangered3",
                            HTH = "orangered4",
                            `High-temperature heat` = "orangered4",
                            `High-temperature\nheat` = "orangered4",
                            LTC = "deepskyblue3",
                            `Low-temperature cooling` = "deepskyblue3",
                            `Low-temperature\ncooling` = "deepskyblue3",
                            IP = "palegreen4",
                            `Information processing` = "palegreen4",
                            `Information\nprocessing` = "palegreen4",
                            L = "gold2",
                            Light = "gold2")
```

```{r stuff-to-exclude, include=FALSE} 
sectors_to_exclude <- RCLabels::make_or_pattern(
  c("Exports", 
    "Exports to World aviation bunkers", 
    "Exports to World marine bunkers", 
    "Losses", 
    "Statistical differences", 
    "Stock changes"))
```



# Introduction

This file investigates questions about
world aggregate mechanical work.
The questions were raised by Tiago Domingos
regarding Figure 7 of the paper
P. E. Brockway, M. K. Heun, Z. Marshall, E. Aramendia, P. Steenwyk, T. Relph, M. Widjanarko,
J. J. Kim, A. Sainju, and J. Irtube. A country-level primary-final-useful (CL-PFU) energy and exergy
database: Overview of its construction and 1971–2020 world-level efficiency results. Environmental
Research: Energy, 1(025005):1–29, jun 2024.

# Data

Gather and save data from the database.
This chunk is not executed unless `eval=TRUE`.

```{r data, eval=FALSE}
conn <- PFUPipelineTools::get_mexerdb_conn(user = "dbcreator")
world_mats <- PFUPipelineTools::pl_filter_collect(
  db_table_name = "PSUTReAllChopAllDsAllGrAll",
  Country == "World",
  conn = conn,
  collect = TRUE)
world_mats |>
  saveRDS(file = "world_mats.rds")
```

# Analysis

Load data from disk.

```{r load}
world_mats <- readRDS(file = "world_mats.rds")
```

Make a graph similar to Figure 7
using pre-aggregated (Grouped) data.
Here, we see that mechanical work (MechW)
is nearly constant, as observed by Tiago.

```{r fig7, fig.width=6, fig.height=3, fig.align='center', fig.retina=2, fig.cap='Remake of Figure 7 using pre-aggregated (i.e., grouped) data.'}
grouped_data <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Grouped", 
                IndustryAggregation == "Grouped") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::select(-R, -U, -U_EIOU, -U_feed, -r_EIOU, -V, -S_units) |>
  dplyr::mutate(
    Y = Y |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude)  |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU") |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |> 
  tidyr::pivot_longer(cols = "Y", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |>
  dplyr::mutate(
    rownames = factor(rownames, levels = c("IP", "L", "HTH", "MTH", "LTH",
                                           "LTC", "HPF", "Prop", "MechW", "HuMech", "AnMech"))

  )
grouped_data |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(limits = c(0, 250)) +
  ggplot2::scale_fill_manual(values = useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme()
```


The Mechanical work numbers in 1971 and 2020 are shown in 
Table \@ref(tab:MechW-table).

```{r MechW-table, results='asis', tab.cap="Mechanical work values in 1971 and 2000."}
# Mechanical work numbers
grouped_data_1971_2020 <- grouped_data |> 
  dplyr::filter(rownames == "MechW", Year %in% c(1971, 2020)) |> 
  dplyr::select(Year, matvals, EnergyType) |> 
  tidyr::pivot_wider(names_from = "EnergyType", values_from = "matvals") |> 
  xtable::xtable(label = "tab:MechW-table", 
                 caption = "Mechanical work values in 1971 and 2000.")
grouped_data_1971_2020 |> 
  print(comment = FALSE, 
        caption.placement = "top", 
        include.rownames = FALSE)
```

```{r perc-increase}
mech_work_1971 <- grouped_data_1971_2020$E[[1]]
mech_work_2020 <- grouped_data_1971_2020$E[[2]]
perc_inc <- (mech_work_2020 - mech_work_1971) / mech_work_1971 * 100
growth_rate <- ((mech_work_2020/mech_work_1971)^(1/(2020-1971)) - 1) * 100
```

Over the 50-year period, 
Mechanical work grows `r sprintf("%.1f", perc_inc)`% for an annual growth rate of
`r sprintf("%.1f", growth_rate)` %/yr.

The actual Mechanical work data are shown in Figure \@ref(fig:mech-work-only).

```{r mech-work-only, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Mechanical work.'}
grouped_data |> 
  dplyr::filter(rownames == "MechW") |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, y = matvals)) + 
  ggplot2::geom_line() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(limits = c(0, NA)) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Mechanical work [EJ]") +
  MKHthemes::xy_theme()
```


Aggregate using specified (the most raw) data to confirm similarity 
of Specified data to Figure 7.

```{r self-aggregation-graph, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Remake of Figure 7 starting from specified data.'}
# Read useful product aggregation map
eu_product_agg_map <- openxlsx2::read_xlsx("aggregation_mapping.xlsx", sheet = "eu_product_aggregation") |> 
  matsbyname::agg_table_to_agg_map(few_colname = "Few", many_colname = "Many")
nouns <- world_mats |> 
  # Filter to the same parameters for Figure 7.
  dplyr::filter(Dataset == "CL-PFU IEA+MW",
                Year >= 1971,
                LastStage == "Useful",
                IncludesNEU == FALSE,
                ProductAggregation == "Specified", 
                IndustryAggregation == "Specified") |>
  dplyr::arrange(Year, EnergyType) |> 
  dplyr::select(-R, -U, -U_EIOU, -U_feed, -r_EIOU, -V, -S_units)

aggregated_nouns <- nouns |> 
  dplyr::mutate(
    Y = Y |> 
      matsbyname::aggregate_pieces_byname(piece = "noun", 
                                          margin = 1,
                                          notation = RCLabels::bracket_notation) |> 
      matsbyname::aggregate_pieces_byname(piece = "noun", 
                                          margin = 2, 
                                          notation = RCLabels::bracket_notation) |> 
      matsbyname::select_cols_byname(remove_pattern = sectors_to_exclude) |> 
      matsbyname::select_rows_byname(remove_pattern = "NEU")
  )
      
grouped <- aggregated_nouns |>       
  dplyr::mutate(
    Y = Y |> 
      matsbyname::aggregate_byname(aggregation_map = list(eu_product_agg_map), 
                                   margin = 1) |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  ) |> 
  tidyr::pivot_longer(cols = "Y", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |> 
  dplyr::mutate(
    rownames = factor(rownames, levels = c("IP", "L", "HTH", "MTH", "LTH",
                                           "LTC", "HPF", "Prop", "MechW", "HuMech", "AnMech"))
    
  )

grouped |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::scale_y_continuous(limits = c(0, 250)) +
  ggplot2::scale_fill_manual(values = useful_carrier_colours) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme()
```


Now focus on Mechanical work, specifically the splits for each portion of mechanical work.
We clearly see that Mechanical work is growing.

```{r mech-work, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Details of mechanical work.'}
mech_work_pattern <- RCLabels::make_or_pattern(eu_product_agg_map$MechW)
mechw_graph <- aggregated_nouns |> 
  dplyr::mutate(
    Y = Y |> 
      matsbyname::select_rows_byname(retain_pattern = mech_work_pattern) |> 
      matsbyname::rowsums_byname() |> 
      matsbyname::hadamardproduct_byname(1e-6) # Convert from TJ to EJ
  )  |> 
  tidyr::pivot_longer(cols = "Y", names_to = "matnames", values_to = "matvals") |>
  matsindf::expand_to_tidy(drop = 0) |> 
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year, 
                                         y = matvals, 
                                         fill = rownames)) + 
  ggplot2::geom_area() + 
  ggplot2::facet_grid(cols = ggplot2::vars(EnergyType)) + 
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  ggplot2::labs(x = NULL, fill = NULL, y = "Useful energy and exergy [EJ]") +
  MKHthemes::xy_theme()
mechw_graph
```

Now, let's plot Mechanical work at the same scale as Figure 7.
Here, we see that Mechanical work looks quite flat, 
although it is growing a bit.

```{r mech-work-full-scale, fig.width=6, fig.height=3, fig.retina=2, fig.align='center', fig.cap='Mechanical work at same scale as Figure 7.'}
mechw_graph +
  ggplot2::scale_y_continuous(limits = c(0, 250)) +
  ggplot2::scale_fill_manual(values = rep(useful_carrier_colours[["MechW"]], 4))
```




# Conclusion
